### 2023-11-22 Presentation/main.tex
Init with Code from Lemire + Paper-Template from LIPIcs + Presentation-Template from s9770652

\documentclass[aspectratio=169]{gpresentation}

\RequirePackage{marvosym}

\graphicspath{{img/}}

\tikzset{
	legend/.style = {font=\boldmath, align=left},
	every path/.style = {line width=.57mm, line cap=round},
}

\subject{Seminar Algorithms for Big Data}
\title{Fast Random Integer Generation in an Interval}
\subtitle{Based on a paper of the same title by Daniel Lemire}
\author{Lukas Geis}
\supervisor{Supervised by Dr Manuel Penschuck}
\date{29th February 2024}
\institute{Algorithm Engineering (Prof. Dr Ulrich Meyer)}

\begin{document}
	\begin{frame}[plain, noframenumbering]
		\titlepage
	\end{frame}

	\begin{frame}{Table of Contents}
		\tableofcontents%[pausesections]
	\end{frame}
\end{document}

### 2023-11-22 FastShuffleExperiments/TOMACS_RCR/cpp/benchmark.h

@@ -1,0 +1,44 @@
+#include <chrono>
+#include <stdint.h>
+#include <assert.h>
+typedef std::chrono::high_resolution_clock Clock;
+
+#define AVG_TIME_NS(test, pre, repeat, size, verbose)                         \
+  do {                                                                         \
+    if (verbose)                                                               \
+      printf("%-60s: ", #test);                                                \
+    assert(repeat >= 5);                                     \
+    fflush(NULL);                                                              \
+    int64_t min_diff = INT64_MAX;                                              \
+    int64_t sum_diff = 0;                                              \
+    for (int i = 0; i < repeat; i++) {                                         \
+      pre;                                                                     \
+      __asm volatile("" ::: /* pretend to clobber */ "memory");                \
+      auto t0 = Clock::now();                                                  \
+      test;                                                                    \
+      auto t1 = Clock::now();                                                  \
+      auto cycles_diff =                                                       \
+          std::chrono::duration_cast<std::chrono::nanoseconds>(t1 - t0)        \
+              .count();                                                        \
+      sum_diff += cycles_diff;                \
+      if (cycles_diff < min_diff)                                              \
+        min_diff = cycles_diff;                                                \
+    }                                                                          \
+    uint64_t S = size;                                                         \
+    if (verbose)                                                               \
+      printf(" %d ns total, ", (int)min_diff);                                 \
+    float cycle_per_op = (min_diff) / (double)S;                               \
+    float avg_cycle_per_op = (sum_diff) / ((double)S * repeat);                \
+    if (verbose)                                                               \
+      printf(" %.2f ns per input key ", cycle_per_op);                         \
+    if (verbose)                                                               \
+      printf("\n");                                                            \
+    bool err = abs(avg_cycle_per_op - cycle_per_op) > 0.01 * avg_cycle_per_op; \
+    double percent = abs(avg_cycle_per_op - cycle_per_op) / avg_cycle_per_op * 100.0; \
+    if(err) {                                                                  \
+      fprintf( stderr, "warning: average %f vs min %f (%f percent) for arrays of size %zu \n", avg_cycle_per_op, cycle_per_op, percent, size); \
+    }                                                                          \
+    if (!verbose)                                                              \
+      printf("  %.2f ", avg_cycle_per_op);                                     \
+    fflush(NULL);                                                              \
+  } while (0)

### 2023-11-22 FastShuffleExperiments/TOMACS_RCR/cpp/shuffle.cpp

@@ -1,0 +1,275 @@
+#include "rangedrand.h"
+#include "shuffle.h"
+#include "benchmark.h"
+#include <algorithm>
+#include <math.h>
+#include <random>
+#include <stdint.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+
+/**
+Sorting stuff
+
+*/
+#define INCLUDESTDSHUFFLE
+int qsort_compare_uint32_t(const void *a, const void *b) {
+  return (*(uint32_t *)a - *(uint32_t *)b);
+}
+
+uint32_t *create_sorted_array(size_t length) {
+  uint32_t *array = (uint32_t *)malloc(length * sizeof(uint32_t));
+  for (size_t i = 0; i < length; i++)
+    array[i] = (uint32_t)pcg32_random();
+  qsort(array, length, sizeof(*array), qsort_compare_uint32_t);
+  return array;
+}
+
+uint32_t *create_random_array(size_t count) {
+  uint32_t *targets = (uint32_t *)malloc(count * sizeof(uint32_t));
+  for (size_t i = 0; i < count; i++)
+    targets[i] = (uint32_t)(pcg32_random() & 0x7FFFFFF);
+  return targets;
+}
+
+// tries to put the array in cache
+void array_cache_prefetch(uint32_t *B, int32_t length) {
+  const int32_t CACHELINESIZE = 64; // 64 bytes per cache line
+  for (int32_t k = 0; k < length; k += CACHELINESIZE / sizeof(uint32_t)) {
+    __builtin_prefetch(B + k);
+  }
+}
+
+#if defined(__x86_64) || defined(__i386__)
+void array_cache_flush(uint32_t *B, int32_t length) {
+  const int32_t CACHELINESIZE = 64; // 64 bytes per cache line
+  for (int32_t k = 0; k < length; k += CACHELINESIZE / sizeof(uint32_t)) {
+    __builtin_ia32_clflush(B + k);
+  }
+}
+#else
+void array_cache_flush(uint32_t *B, int32_t length) { (void) B; }
+
+#endif
+
+int sortAndCompare(uint32_t *shuf, uint32_t *orig, uint32_t size) {
+  qsort(shuf, size, sizeof(uint32_t), qsort_compare_uint32_t);
+  qsort(orig, size, sizeof(uint32_t), qsort_compare_uint32_t);
+  for (uint32_t k = 0; k < size; ++k)
[…]

### 2023-11-22 FastShuffleExperiments/cpp/chisqr.h

@@ -1,0 +1,179 @@
+
+/*
+    Implementation of the Chi-Square Distribution &
+        Incomplete Gamma Function in C
+
+        Written By Jacob Wells
+        July 31, 2012
+    Based on the formulas found here:
+
+    Wikipedia - Incomplete Gamma Function -> Evaluation formulae -> Connection
+   with Kummer's confluent hypergeometric function
+    http://en.wikipedia.org/wiki/Regularized_Gamma_function#Connection_with_Kummer.27s_confluent_hypergeometric_function
+
+    Wikipedia - Chi-squared Distribution -> Cumulative distribution function
+    http://en.wikipedia.org/wiki/Chi-squared_distribution#Cumulative_distribution_function
+
+    These functions are placed in the Public Domain, and may be used by anyone,
+   anywhere, for any reason, absolutely free of charge.
+
+*/
+
+#include <math.h>
+#include <stdio.h>
+/** gamma function
+ * taken from http://www.crbond.com/math.htm (Gamma function in C/C++ for real
+ * arguments, ported from Zhang and Jin)
+ * returns 1e308 if argument is a negative integer or 0 or if argument exceeds
+ * 171.
+ * @param x     argument
+ * @return      Gamma(argument)
+ */
+static double gamma0(double x) {
+  /* gamma function.
+  Algorithms and coefficient values from "Computation of Special
+  Functions", Zhang and Jin, John Wiley and Sons, 1996.
+  (C) 2003, C. Bond. All rights reserved.
+  taken from http://www.crbond.com/math.htm */
+  int i, k, m;
+  double ga, gr, r = 1.0, z;
+
+  static double g[] = {1.0,
+                       0.5772156649015329,
+                       -0.6558780715202538,
+                       -0.420026350340952e-1,
+                       0.1665386113822915,
+                       -0.421977345555443e-1,
+                       -0.9621971527877e-2,
+                       0.7218943246663e-2,
+                       -0.11651675918591e-2,
+                       -0.2152416741149e-3,
+                       0.1280502823882e-3,
+                       -0.201348547807e-4,
+                       -0.12504934821e-5,
+                       0.1133027232e-5,
[…]

### 2023-11-22 FastShuffleExperiments/cpp/justone.cpp

@@ -1,0 +1,82 @@
+
+#include "shuffle.h"
+#include "benchmark.h"
+#include <algorithm>
+#include <math.h>
+#include <random>
+#include <stdint.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+
+uint32_t *create_array(size_t count) {
+  uint32_t *targets = (uint32_t *)malloc(count * sizeof(uint32_t));
+  for (size_t i = 0; i < count; i++)
+    targets[i] = i;
+  return targets;
+}
+
+void ShuffleBenchmark(size_t size, int mode, bool verbose) {
+  if (verbose) {
+
+    printf(" %s\n", __PRETTY_FUNCTION__);
+    printf("Shuffling arrays of size %zu \n", size);
+    printf("Time reported in number of cycles per array element.\n");
+    printf("Tests assume that array is in cache as much as possible.\n");
+  } else {
+    printf("%zu or %f MB ", size, size * 4.0/(1024.0 * 1024.0));
+  }
+  uint32_t *testvalues = create_array(size);
+  switch(mode) {
+    case 1:
+    printf("shuffle_go32\n");
+    shuffle_go32<lehmer64_32>(testvalues, size);
+    break;
+    case 2:
+    printf("shuffle_java32\n");
+    shuffle_java32<lehmer64_32>(testvalues, size);
+    break;
+    case 3:
+    printf("shuffle_nearlydivisionless32\n");
+    shuffle_nearlydivisionless32<lehmer64_32>(testvalues, size);
+    break;
+    case 4:
+    printf("shuffle_floatmult32\n");
+    shuffle_floatmult32<lehmer64_32>(testvalues, size);
+    break;
+    case 5:
+    printf("shuffle_go64\n");
+    shuffle_go64<lehmer64>(testvalues, size);
+    break;
+    case 6:
+    printf("shuffle_java64\n");
+    shuffle_java64<lehmer64>(testvalues, size);
+    break;
+    case 7:
+    printf("shuffle_nearlydivisionless64\n");
+    shuffle_nearlydivisionless64<lehmer64>(testvalues, size);
+    break;
+    case 8:
+    printf("shuffle_floatmult64\n");
+    shuffle_floatmult64<lehmer64>(testvalues, size);
+    break;
+    default:
+    break;
+  }
+
+  printf("%u \n", testvalues[0]);
+  free(testvalues);
+}
+
+
+
[…]

### 2023-11-22 FastShuffleExperiments/cpp/lehmer64.h

@@ -1,0 +1,25 @@
+#ifndef LEHMER64_H
+#define LEHMER64_H
+
+#include "splitmix64.h"
+
+static __uint128_t g_lehmer64_state;
+
+/**
+* D. H. Lehmer, Mathematical methods in large-scale computing units.
+* Proceedings of a Second Symposium on Large Scale Digital Calculating
+* Machinery;
+* Annals of the Computation Laboratory, Harvard Univ. 26 (1951), pp. 141-146.
+*/
+
+static inline void lehmer64_seed(uint64_t seed) {
+  g_lehmer64_state = (((__uint128_t)splitmix64_stateless(seed)) << 64) +
+                     splitmix64_stateless(seed + 1);
+}
+
+static inline uint64_t lehmer64() {
+  g_lehmer64_state *= UINT64_C(0xda942042e4dd58b5);
+  return g_lehmer64_state >> 64;
+}
+
+#endif

### 2023-11-22 FastShuffleExperiments/cpp/rangedrand.h

@@ -1,0 +1,188 @@
+#ifndef RANGEDRAND_H
+#define RANGEDRAND_H
+#include <random>
+#include <stdint.h>
+
+#include "lehmer64.h"
+#include "pcg.h"
+#include "splitmix64.h"
+#include "xorshift128plus.h"
+
+static void setseed(uint64_t seed) {
+  xorshift128plus_seed(seed);
+  splitmix64_seed(seed);
+  lehmer64_seed(seed);
+}
+
+static inline uint32_t lehmer64_32(void) { return (uint32_t)lehmer64(); }
+
+static inline uint32_t xorshift128plus_32(void) {
+  return (uint32_t)xorshift128plus();
+}
+
+std::mt19937 mersenne;
+
+static inline uint32_t twister32(void) { return (uint32_t)mersenne(); }
+
+typedef uint64_t (*randfnc64)(void);
+
+typedef uint32_t (*randfnc32)(void);
+
+// wrapper to satisfy fancy C++
+template <randfnc32 RandomBitGenerator>
+struct UniformRandomBitGenerator32Struct {
+  typedef uint32_t result_type;
+  static constexpr result_type min() { return 0; }
+  static constexpr result_type max() { return UINT32_MAX; }
+  result_type operator()() { return RandomBitGenerator(); }
+};
+
+// wrapper to satisfy fancy C++
+template <randfnc64 RandomBitGenerator>
+struct UniformRandomBitGenerator64Struct {
+  typedef uint64_t result_type;
+  static constexpr result_type min() { return 0; }
+  static constexpr result_type max() { return UINT64_MAX; }
+  result_type operator()() { return RandomBitGenerator(); }
+};
+
+// return value in [0,bound)
+template <randfnc32 RandomBitGenerator>
+static inline uint32_t naive_random_bounded32(uint32_t bound) {
+  return RandomBitGenerator() % bound;
+}
+
+template <randfnc64 RandomBitGenerator>
+static inline uint64_t naive_random_bounded64(uint64_t bound) {
+  return RandomBitGenerator() % bound;
+}
+
+// return value in [0,bound)
+// as per the PCG implementation, uses two 32-bit divisions
+template <randfnc32 RandomBitGenerator>
+static inline uint32_t random_bounded32(uint32_t bound) {
+  uint32_t threshold = (~bound + 1) % bound; // -bound % bound
+  for (;;) {
+    uint32_t r = RandomBitGenerator();
+    if (r >= threshold)
+      return r % bound;
+  }
+}
+
[…]

### 2023-11-22 FastShuffleExperiments/cpp/shuffle.cpp

@@ -1,0 +1,273 @@
+#include "rangedrand.h"
+#include "shuffle.h"
+#include "benchmark.h"
+#include <algorithm>
+#include <math.h>
+#include <random>
+#include <stdint.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+
+/**
+Sorting stuff
+
+*/
+#define INCLUDESTDSHUFFLE
+int qsort_compare_uint32_t(const void *a, const void *b) {
+  return (*(uint32_t *)a - *(uint32_t *)b);
+}
+
+uint32_t *create_sorted_array(size_t length) {
+  uint32_t *array = (uint32_t *)malloc(length * sizeof(uint32_t));
+  for (size_t i = 0; i < length; i++)
+    array[i] = (uint32_t)pcg32_random();
+  qsort(array, length, sizeof(*array), qsort_compare_uint32_t);
+  return array;
+}
+
+uint32_t *create_random_array(size_t count) {
+  uint32_t *targets = (uint32_t *)malloc(count * sizeof(uint32_t));
+  for (size_t i = 0; i < count; i++)
+    targets[i] = (uint32_t)(pcg32_random() & 0x7FFFFFF);
+  return targets;
+}
+
+// tries to put the array in cache
+void array_cache_prefetch(uint32_t *B, int32_t length) {
+  const int32_t CACHELINESIZE = 64; // 64 bytes per cache line
+  for (int32_t k = 0; k < length; k += CACHELINESIZE / sizeof(uint32_t)) {
+    __builtin_prefetch(B + k);
+  }
+}
+
+#if defined(__x86_64) || defined(__i386__)
+void array_cache_flush(uint32_t *B, int32_t length) {
+  const int32_t CACHELINESIZE = 64; // 64 bytes per cache line
+  for (int32_t k = 0; k < length; k += CACHELINESIZE / sizeof(uint32_t)) {
+    __builtin_ia32_clflush(B + k);
+  }
+}
+#else
+void array_cache_flush(uint32_t *B, int32_t length) { (void) B; }
+
+#endif
+
+int sortAndCompare(uint32_t *shuf, uint32_t *orig, uint32_t size) {
+  qsort(shuf, size, sizeof(uint32_t), qsort_compare_uint32_t);
+  qsort(orig, size, sizeof(uint32_t), qsort_compare_uint32_t);
+  for (uint32_t k = 0; k < size; ++k)
[…]

### 2023-11-22 FastShuffleExperiments/cpp/sillyreservoir.cpp

@@ -1,0 +1,108 @@
+
+#include "sillyreservoir.h"
+#include "benchmark.h"
+#include <algorithm>
+#include <math.h>
+#include <random>
+#include <stdint.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+
+#define INCLUDEFLOAT
+
+template <randfnc32 rfnc32> void ReservoirBenchmark32(size_t capacity, size_t size, bool verbose) {
+  if (verbose) {
+
+    printf(" %s\n", __PRETTY_FUNCTION__);
+    printf("Time reported in number of cycles per array element.\n");
+    printf("Tests assume that array is in cache as much as possible.\n");
+  } else {
+    printf("%zu ", size);
+  }
+  int repeat = 10;
+  if (size < 1000000)
+    repeat = 50;
+  if (size > 10000000)
+    repeat = 5;
+  uint32_t *reservoir = (uint32_t *) malloc(capacity * sizeof(uint32_t));
+
+  BEST_TIME_NS(sillyreservoirsampling_go32<rfnc32>(reservoir, capacity, size),
+               , repeat, size, verbose);
+
+  BEST_TIME_NS(sillyreservoirsampling_java32<rfnc32>(reservoir, capacity, size),
+               , repeat, size, verbose);
+
+  BEST_TIME_NS(sillyreservoirsampling_nearlydivisionless32<rfnc32>(reservoir, capacity, size),
+               , repeat, size, verbose);
+
+#ifdef INCLUDEFLOAT
+  BEST_TIME_NS(sillyreservoirsampling_floatmult32<rfnc32>(reservoir, capacity, size),
+               , repeat, size, verbose);
+
+#endif
+
+  printf("        %d ", reservoir[0]);
+  free(reservoir);
+  printf("\n");
+}
+
+template <randfnc64 rfnc64> void ReservoirBenchmark64(size_t capacity, size_t size, bool verbose) {
+  if (verbose) {
+    printf(" %s\n", __PRETTY_FUNCTION__);
+    printf("Time reported in number of ns per array element.\n");
+    printf("Tests assume that array is in cache as much as possible.\n");
+  } else {
+    printf("%zu ", size);
+  }
+  int repeat = 10;
+  if (size < 1000000)
+    repeat = 50;
+  if (size > 10000000)
+    repeat = 5;
+  uint32_t *reservoir = (uint32_t *) malloc(capacity * sizeof(uint32_t));
+
[…]

### 2023-11-22 FastShuffleExperiments/cpp/stattests.cpp

@@ -1,0 +1,104 @@
+#include "chisqr.h"
+#include "rangedrand.h"
+#include <cassert>
+#include <math.h>
+#include <stdint.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <vector>
+
+typedef uint32_t (*rangedfnc32)(uint32_t);
+typedef uint64_t (*rangedfnc64)(uint64_t);
+
+template <rangedfnc32 fnc>
+bool isuniform32(size_t N, double &ret, bool verbose = true,
+                 double tol = 0.01) {
+  std::vector<double> counts(N);
+  size_t volume = 10000 * N;
+  for (size_t i = 0; i < volume; i++) {
+    counts[fnc(N)] += 1.0;
+  }
+  double expected = (double)volume / N;
+  double sum = 0;
+  for (size_t i = 0; i < N; i++) {
+    double x = counts[i] - expected;
+    sum += x * x / expected;
+  }
+  ret = chisqr(N - 1, sum);
+  if (ret < tol) {
+    if (verbose)
+      printf(" %s fails with %f [%f] N = %zu \n", __PRETTY_FUNCTION__, ret, sum,
+             N);
+  }
+  return ret >= tol;
+}
+
+template <rangedfnc64 fnc>
+bool isuniform64(size_t N, double &ret, bool verbose = true,
+                 double tol = 0.01) {
+  std::vector<double> counts(N);
+  size_t volume = 10000 * N;
+  for (size_t i = 0; i < volume; i++) {
+    counts[fnc(N)] += 1.0;
+  }
+  double expected = (double)volume / N;
+  double sum = 0;
+  for (size_t i = 0; i < N; i++) {
+    double x = counts[i] - expected;
+    sum += x * x / expected;
+  }
+  ret = chisqr(N - 1, sum);
+  if (ret < tol) {
+    if (verbose)
+      printf(" %s fails with %f [%f] N = %zu \n", __PRETTY_FUNCTION__, ret, sum,
+             N);
+  }
+  return ret >= tol;
+}
+
+uint32_t bad(uint32_t x) { return x - 1; }
+
+uint32_t badbutbetter(uint32_t x) { return rand() % (x - 1); }
+uint32_t notgreat(uint32_t x) { return rand() % x; }
+
+int main() {
+  setseed(1234);
+  size_t success = 0;
+  size_t total = 0;
+  for (size_t N = 2; N <= 300; N += 1) {
[…]

### 2023-11-22 FastShuffleExperiments/extra/go/shuffle.go

@@ -1,0 +1,101 @@
+package shuffle
+
+import "math"
+import (
+	"github.com/dgryski/go-pcgr"
+)
+
+// return a pseudo-random number in [0,r), using the go-like approach, uses the dgryski' PCG for random bits
+func Random_bounded_go(r uint32, pcg *pcgr.Rand) uint32 {
+	max := uint32(0xFFFFFFFF) % r
+	v := pcg.Next()
+	for v <= max {
+		v = pcg.Next()
+	}
+	return v % r
+}
+
+// return a pseudo-random number in [0,r), using the java-like approach, uses the dgryski' PCG for random bits
+func Random_bounded_java(r uint32, pcg *pcgr.Rand) uint32 {
+	v := pcg.Next()
+	candidate := v % r
+	for v-candidate > math.MaxUint32-r+uint32(1) {
+		v = pcg.Next()
+		candidate = v % r
+	}
+	return candidate
+}
+
+// return a pseudo-random number in [0,r), avoiding divisions, uses the dgryski' PCG for random bits
+func Random_bounded_divisionless(r uint32, pcg *pcgr.Rand) uint32 {
+	random32bit := pcg.Next()
+	multiresult := uint64(random32bit) * uint64(r)
+	leftover := uint32(multiresult)
+	if leftover < r {
+		threshold := uint32(-r) % r
+		for leftover < threshold {
+			random32bit = pcg.Next()
+			multiresult = uint64(random32bit) * uint64(r)
+			leftover = uint32(multiresult)
+		}
+	}
+	return uint32(multiresult >> 32) // [0, r)
+}
+
+func shuffle_divisionless(array []int, p *pcgr.Rand) {
+	r := len(array)
+	for i := r; i > 0; i-- {
+		//		idx := Random_bounded_divisionless(uint32(i), p)
+		random32bit := p.Next()
+		multiresult := uint64(random32bit) * uint64(i)
+		leftover := uint32(multiresult)
+		if leftover < uint32(i) {
+			threshold := uint32(-i) % uint32(i)
+			for leftover < threshold {
+				random32bit = p.Next()
+				multiresult = uint64(random32bit) * uint64(i)
+				leftover = uint32(multiresult)
+			}
+		}
+		idx := uint32(multiresult >> 32) // [0, r)
+		array[idx], array[i-1] = array[i-1], array[idx]
+	}
+}
+
[…]

### 2023-11-22 FastShuffleExperiments/extra/go/shuffle_test.go

@@ -1,0 +1,70 @@
+package shuffle
+
+import "testing"
+import (
+	"github.com/dgryski/go-pcgr"
+)
+
+func makeRandomArray(r int, pcg *pcgr.Rand) []int {
+	array := make([]int, r, r)
+	for i := 0; i < r; i++ {
+		array[i] = int(pcg.Next() & 0x7FFFFFFF)
+	}
+	return array
+}
+
+func benchmark_pcg(b *testing.B, r int) {
+	b.StopTimer()
+	p := pcgr.Rand{0x0ddc0ffeebadf00d, 0xcafebabe}
+	array := makeRandomArray(r, &p)
+	b.StartTimer()
+	for j := 0; j < b.N; j++ {
+		shuffle_pcg(array, &p)
+	}
+}
+
+func BenchmarkPcg1000(b *testing.B) {
+	benchmark_pcg(b, 1000)
+}
+
+func benchmark_divisionless(b *testing.B, r int) {
+	b.StopTimer()
+	p := pcgr.Rand{0x0ddc0ffeebadf00d, 0xcafebabe}
+	array := makeRandomArray(r, &p)
+	b.StartTimer()
+	for j := 0; j < b.N; j++ {
+		shuffle_divisionless(array, &p)
+	}
+}
+
+func BenchmarkDivisionless1000(b *testing.B) {
+	benchmark_divisionless(b, 1000)
+}
+
+func benchmark_go(b *testing.B, r int) {
+	b.StopTimer()
+	p := pcgr.Rand{0x0ddc0ffeebadf00d, 0xcafebabe}
+	array := makeRandomArray(r, &p)
+	b.StartTimer()
+	for j := 0; j < b.N; j++ {
+		shuffle_go(array, &p)
+	}
+}
+
+func BenchmarkGo1000(b *testing.B) {
+	benchmark_go(b, 1000)
+}
+
+func benchmark_java(b *testing.B, r int) {
+	b.StopTimer()
+	p := pcgr.Rand{0x0ddc0ffeebadf00d, 0xcafebabe}
+	array := makeRandomArray(r, &p)
+	b.StartTimer()
+	for j := 0; j < b.N; j++ {
+		shuffle_java(array, &p)
+	}
+}
+
+func BenchmarkJava1000(b *testing.B) {
+	benchmark_java(b, 1000)
+}

### 2023-11-22 FastShuffleExperiments/extra/java/src/main/java/me/lemire/microbenchmarks/algorithms/Shuffle.java

@@ -1,0 +1,186 @@
+package me.lemire.microbenchmarks.algorithms;
+
+import java.util.Arrays;
+import java.util.Random;
+import java.util.concurrent.ThreadLocalRandom;
+import java.util.concurrent.TimeUnit;
+
+import org.openjdk.jmh.annotations.*;
+import org.openjdk.jmh.runner.Runner;
+import org.openjdk.jmh.runner.RunnerException;
+import org.openjdk.jmh.runner.options.Options;
+import org.openjdk.jmh.runner.options.OptionsBuilder;
+
+@State(Scope.Benchmark)
+@BenchmarkMode(Mode.AverageTime)
+@OutputTimeUnit(TimeUnit.NANOSECONDS)
+public class Shuffle {
+
+    private static void swap(int[] arr, int i, int j) {
+        int tmp = arr[i];
+        arr[i] = arr[j];
+        arr[j] = tmp;
+    }
+
+    private static int nextPositiveInteger(Random rnd) {
+        return rnd.nextInt() & 0x7fffffff;
+    }
+
+    private static void shuffle_java(int arr[], Random rnd) {
+        int size = arr.length;
+        // Shuffle array
+        for (int i = size; i > 1; i--)
+            swap(arr, i - 1, rnd.nextInt(i));
+    }
+
+    private static int ranged_random_go(int  range,  Random rnd) {
+        int bits;
+        int t = Integer.MAX_VALUE % range;
+        do {
+            bits = nextPositiveInteger(rnd);
+        } while(bits <= t);
+        return bits % range;
+    }
+
+    private static void shuffle_go(int arr[], Random rnd) {
+        int size = arr.length;
+        // Shuffle array
+        for (int i = size; i > 1; i--)
+            swap(arr, i - 1, ranged_random_go(i,rnd));
+    }
+
+
+
+    private static int ranged_random_openbsd(int  range,  Random rnd) {
+        int threshold = (Integer.MAX_VALUE - range + 1) % range;
+        for (;;) {
+            int u = nextPositiveInteger(rnd);
+            if (u >= threshold)
+                return u % range;
+        }
+    }
+
+    private static void shuffle_openbsd(int arr[], Random rnd) {
+        int size = arr.length;
+        // Shuffle array
+        for (int i = size; i > 1; i--)
+            swap(arr, i - 1, ranged_random_openbsd(i,rnd));
+    }
+
+    private static int ranged_random_javalike(int  range,  Random rnd) {
[…]

### 2023-11-28 Paper/main.tex
Motivation

\documentclass[a4paper, UKenglish, cleveref, autoref, thm-restate]{lipics-v2021}

\bibliographystyle{plainurl}

\title{Fast Random Integer Generation in an Interval}

\author{Lukas Geis}{Goethe Universität Frankfurt am Main}{s2949316@stud.uni-frankfurt.de}{}{}

\authorrunning{L. Geis}

\Copyright{Lukas Geis}

\ccsdesc[100]{Mathematics of computing $\rightarrow$ Random number generation}

\keywords{Random Number Generation, Rejection Method, Randomized Algorithms, Algorithm Engineering}

\category{} 
\relatedversion{} 
\supplement{Sources of implementations and experiments can be accessed at \url{https://github.com/lukasgeis/FRIGiaI-Review}}
\acknowledgements{}
\nolinenumbers

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Packages
\usepackage{mathtools}

\usepackage[vlined,ruled,linesnumbered]{algorithm2e}
\DontPrintSemicolon
\SetKwInOut{Require}{Require}

\usepackage{hyperref}

\begin{document}

\maketitle

%%%%% Abstract
\begin{abstract}
    Lorem ipsum dolor sit amet, consectetur adipiscing elit. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti. 
\end{abstract}

%%%%% Introduction
\section{Introduction}
\subsection{Motivation}
The study of randomized algorithms goes as far back as the 1910s~\cite{MathComp}.
While initially small, this study is now scientifically pursued in almost all parts of computer science.
Not only in theory, but also in practice, randomized algorithms gain even more importance today to not only solve specific algorithmic problems but also to create large, scalable, and unbiased data sets that can be used in a variety of fields.
\\
The core of such algorithms is most often the pseudo-random number generator (PRNG).
There are many efficient PRNGs such as XorShift~\cite{XS}, MersenneTwister~\cite{MT}, linear congruential generators~\cite{LinConGen, MINSTD, LM}, and younger generators such as the PCG-Family~\cite{PCG} and ChaCha~\cite{Cha}.
These generators produce $32$-bit or $64$-bit words which we can interpret as unsigned integers in $[0,2^{32})$ and $[0,2^{64})$ respectively.
However, for most applications we do not want to have such a large range of possible random values but instead want to confine the range to a given interval $[a,b]$.
This can not only be used to uniformly sample an element from an array but also for more complex algorithms
\begin{itemize}
    \item The Fisher-Yates random shuffle~\cite{FY} (see \hyperref[alg:fy]{Algorithm 1}) uniformly permutes an array of $n$ elements in time $\Theta(n)$. 
    We have to draw a random integer in an interval for every shuffled element.
    \item A random graph model describes a distribution over a (parametrized) family of graphs. 
    There exist an uncountable number of such models and almost all of them require some sort of integer sampling in an interval. 
    For example, the $\mathcal{G}(n,m)$ model~\cite{GNM} uniformly samples a (directed) graph with $n$ nodes and $m$ edges. 
    A possible algorithm for this problem is Reservoir-Sampling~\cite{Reservoir}(see \hyperref[alg:gnm]{Algorithm 2}) which is a general technique that can sample $k$ elements without replacement from an input stream. 
    Another model is the BA model~\cite{BABook,BA} which iteratively builds the graph by adding nodes with edges to previously added nodes (see \hyperref[alg:ba]{Algorithm 3}). 
    Both models require sampling of an integer in an interval for each of its edges (or more).
    \item An Alias-Table~\cite{Alias} can be used to simulate static discrete distributions in practice. 
    After creating and initializing the table, we can sample in (expected) time $\Theta(1)$ by drawing a uniform random row and offset in each round. 
    Hence, we need to sample an integer uniformly at random in $[0,n)$ where $n$ is the number of rows in the table.
\end{itemize}

%%% BEGIN Fisher-Yates-Shuffle
\begin{algorithm}[!htb] \label{alg:fy}
    \caption{Fisher-Yates-Shuffle: uniformly shuffle an array of size $n$}
    \Require{source of uniformly-distrbuted random integers in $[0,i]$ for parameter $i$}
    \Require{array $A$ of size $n$}
    \For{$i = n - 1,\ldots,1$}{
        $j \leftarrow$ random integer in $[0,i]$\;
        swap $A[i]$ and $A[j]$\;
    }
\end{algorithm}
%%% END Fisher-Yates-Shuffle

%%% BEGIN G(n,m) Sampling
\begin{algorithm}[!htb] \label{alg:gnm}
    \caption{Sampling of (directed) $\mathcal{G}(n,m)$ graphs using Reservoir Sampling}
    \Require{source of uniformly-distrbuted random integers in $[0,i]$ for parameter $i$}
    \Require{number of nodes $n$ and number of edges $m \leq n^2$}
    $E \leftarrow$ array (of edges) of size $m$\;
    \For{$i = 0,\ldots,m - 1$}{
        $E[i] \leftarrow \left(i \div n, i \mod n\right)$\;
    }
    \For{$i = m,\ldots,n^2$}{
        $j \leftarrow$ random integer in $[0,i]$\;
        \If{$j < m$}{
            $E[j] \leftarrow \left(i \div n, i \mod n\right)$\;
        }
    }
    \KwRet{$E$}\;
\end{algorithm}
%%% END G(n,m) Sampling

%%% BEGIN BA-Sampling
\begin{algorithm}[!htb] \label{alg:ba}
    \caption{Preferential-Attachment-Generator \cite{BASampling}: sample BA-graphs in linear time}
    \Require{source of uniformly-distrbuted random integers in $[0,i]$ for parameter $i$}
    \Require{inital edge-list $E_0$ with $n_0$ nodes and $m_0 \coloneqq |E_0|$ edges, number of additional nodes $N$, neighbor-parameter $\nu$}
    $M \leftarrow$ array of size $2\left(m_0 + \nu N\right)$\;
    \For{$i = 0,\ldots,m_0 - 1$}{
        $\left(M[2i], M[2i + 1]\right) \leftarrow E_0[i]$\;
    }
    \For{$i = 0,\ldots,N - 1$}{
        \For{$j = 1,\ldots,\nu$}{
            $k \leftarrow$ random integer in $[0,2(m_0 + i\nu) - 1]$\;
            $M[2(m_0 + i\nu)] \leftarrow M[k]$\;
            $M[2(m_0 + i\nu) + 1] \leftarrow n_0 + i$\;
        }
    }
    $E \leftarrow$ array (of edges) of size $\left(m_0 + \nu N\right)$\;
    \For{$i = 0,\ldots,m_0 + \nu N - 1$}{
        $E[i] \leftarrow (M[2i], M[2i + 1])$\;
    }
    \KwRet{$E$}\;
\end{algorithm}
%%% END BA-Sampling

\subsection{Preliminaries}

\subsection{Related Work}

\subsection{Structure}

%%%%% Techniques
\section{Existing Techniques}
\subsection{Biased}

\subsection{Unbiased}

%%%%% Lemire
\section{Nearly Divisionless Unbiased Sampling}

%%%%% Experiments
\section{Experiments}
\subsection{Shuffling-Experiments}

\subsection{In-Memory-Processing-Experiments}

%%%%% Conclusion
\section{Conclusion}

\bibliography{refs}

\end{document}

### 2023-11-29 Paper/main.tex
Parts of 1

@@
 \usepackage{mathtools}
+\usepackage{nicefrac}
+\newcommand{\Mod}[1]{\ \mathrm{mod}\ #1}
 
@@
 \subsection{Preliminaries}
+We denote by $x \div y$ the integer division operation, namely $\lfloor\nicefrac{x}{y}\rfloor$, and the remainder operation by $x \Mod y \equiv x - (x \div y)y$. 
+Divisions by a power of two can be efficiently implemented by Bit-\textsc{RightShift} denoted by $\gg$: $x \div 2^W \equiv x \gg W$.
+Similarly, we can efficiently compute the remainder of a division py a power of two by a logical \textsc{And}-operation, denoted by $\&$: $x \Mod 2^W \equiv x \And (2^W - 1)$.
+Typically, PRNGs generate uniform random numbers in an interval $[0,2^W)$ where $W \in \{32,64\}$, namely a uniform random $W$-bit number.
+
+Since we can map every integer in an interval $[a,b)$ to the interval $[0,b - a)$ by subtracting $a$, our goal is to generate a uniform random integer in an interval $[0,s)$ for a given $s$.
+If we map all integers $x$ in $[0,2^W)$ with the function $x \Mod s$, we get the following sequence:
+{\Large
+\[
+    \overbrace{\underbrace{\overbrace{0,1,\ldots,s - 1}^{\text{$s$ values}},\overbrace{0,1,\ldots,s - 1}^{\text{$s$ values}},\ldots,\overbrace{0,1,\ldots,s - 1}^{\text{$s$ values}}}_{\text{$\left(2^W \div s\right) \cdot s$ values}},\underbrace{0,1,\ldots,\left(2^W \Mod s\right) - 1}_{\text{$2^W \Mod s$ values}}}^{\text{$2^W$ values}}
+\]
+}%
+This motivates the following lemma:
+\begin{lemma}[Remainder Multiples]
+    For given integers $a,b,s > 0$ with $a < b$, there exist exactly $(b - a) \div s$ integers $x \in [a,b)$ for every $0 \leq r < s$ such that $x \Mod s \equiv r$ whenever $s$ divides $b - a$.
+\end{lemma}
 
@@
 \subsection{Structure}
+We present and analyse existing techniques for uniformly sampling random integers in an interval $[0,s)$ in \hyperref[sec:2]{Section 2}.
+In \hyperref[sec:3]{Section 3}, we present the main contribution of this paper: a nearly-divisionless algorithm to sampler uniform random integers in an interval $[0,s)$.
+In \hyperref[sec:4]{Section 4}, we benchmark this new technique in two experiments - shuffling and performance in the novel \textsc{Upmem-Pim} architecture - and \hyperref[sec:5]{Section 5} contains a summary of all results.
 
 %%%%% Techniques
-\section{Existing Techniques}
+\section{Existing Techniques}\label{sec:2}
 \subsection{Biased}
@@
 %%%%% Lemire
-\section{Nearly Divisionless Unbiased Sampling}
+\section{Nearly Divisionless Unbiased Sampling}\label{sec:3}
 
 %%%%% Experiments
-\section{Experiments}
+\section{Experiments}\label{sec:4}
 \subsection{Shuffling-Experiments}
@@
 %%%%% Conclusion
-\section{Conclusion}
+\section{Conclusion}\label{sec:5}
 

### 2023-12-02 Paper/main.tex
restructure report a bit
...

@@
 
-Since we can map every integer in an interval $[a,b)$ to the interval $[0,b - a)$ by subtracting $a$, our goal is to generate a uniform random integer in an interval $[0,s)$ for a given $s$.
+Since we can map every integer in an interval $[a,b)$ to the interval $[0,b - a)$ by subtracting $a$, we want to generate a uniform random integer in an interval $[0,s)$ for a given $s$.
 If we map all integers $x$ in $[0,2^W)$ with the function $x \Mod s$, we get the following sequence:
@@
 In \hyperref[sec:3]{Section 3}, we present the main contribution of this paper: a nearly-divisionless algorithm to sampler uniform random integers in an interval $[0,s)$.
-In \hyperref[sec:4]{Section 4}, we benchmark this new technique in two experiments - shuffling and performance in the novel \textsc{Upmem-Pim} architecture - and \hyperref[sec:5]{Section 5} contains a summary of all results.
+In \hyperref[sec:4]{Section 4}, we provide a short summary.
 
@@
 
+\subsubsection{Modulo Reduction}
+
+\subsubsection{Multiply-and-Shift}
+
+\subsubsection{Floating-Point Conversion}
+
 \subsection{Unbiased}
+
+\subsubsection{OpenBSD}
+
+\subsubsection{Java}
+
+\subsubsection{Fast-Dice-Roller}
+
+\subsubsection{Bitmask}
 
@@
 
-%%%%% Experiments
-\section{Experiments}\label{sec:4}
-\subsection{Shuffling-Experiments}
-
-\subsection{In-Memory-Processing-Experiments}
-
 %%%%% Conclusion
-\section{Conclusion}\label{sec:5}
+\section{Conclusion}\label{sec:4}
 

### 2023-12-03 Paper/main.tex
Section 2

--- 2023-11-28 Paper/main.tex
@@
     \caption{Fisher-Yates-Shuffle: uniformly shuffle an array of size $n$}
-    \Require{source of uniformly-distrbuted random integers in $[0,i]$ for parameter $i$}
+    \Require{source of uniformly-distributed random integers in $[0,i]$ for parameter $i$}
     \Require{array $A$ of size $n$}
@@
     \caption{Sampling of (directed) $\mathcal{G}(n,m)$ graphs using Reservoir Sampling}
-    \Require{source of uniformly-distrbuted random integers in $[0,i]$ for parameter $i$}
+    \Require{source of uniformly-distributed random integers in $[0,i]$ for parameter $i$}
     \Require{number of nodes $n$ and number of edges $m \leq n^2$}
@@
     \caption{Preferential-Attachment-Generator \cite{BASampling}: sample BA-graphs in linear time}
-    \Require{source of uniformly-distrbuted random integers in $[0,i]$ for parameter $i$}
+    \Require{source of uniformly-distributed random integers in $[0,i]$ for parameter $i$}
     \Require{inital edge-list $E_0$ with $n_0$ nodes and $m_0 \coloneqq |E_0|$ edges, number of additional nodes $N$, neighbor-parameter $\nu$}
@@
 
-\subsection{Preliminaries}
+\subsection{Preliminaries}\label{sec:1.2}
+We denote by $x \div y$ the integer division operation, namely $\lfloor\nicefrac{x}{y}\rfloor$, and the remainder operation by $x \Mod y \equiv x - (x \div y)y$. 
+Divisions by a power of two can be efficiently implemented by Bit-\textsc{RightShift} denoted by $\gg$: $x \div 2^W \equiv x \gg W$.
+Similarly, we denote a Bit-\textsc{LeftShift} by $\ll$.
+Additionally, we can efficiently compute the remainder of a division py a power of two by a logical \textsc{And}-operation, denoted by $\&$: $x \Mod 2^W \equiv x \And (2^W - 1)$.
+Typically, PRNGs generate uniform random numbers in an interval $[0,2^W)$ where $W \in \{32,64\}$, namely a uniform random $W$-bit number.
 
-\subsection{Related Work}
+Since we can map every integer in an interval $[a,b)$ to the interval $[0,b - a)$ by subtracting $a$, we want to generate a uniform random integer in an interval $[0,n)$ for a given $n$.
+If we map all integers $x$ in $[0,2^W)$ with the function $x \Mod n$, we get the following sequence:
+{\large
+\begin{align}\label{eq:1}
+    \overbrace{\underbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{$\left(2^W \div n\right) \cdot n$ values}},\underbrace{0,1,\ldots,\left(2^W \Mod n\right) - 1}_{\text{$2^W \Mod n$ values}}}^{\text{$2^W$ values}}
+\end{align}
+}%
+This motivates the following lemma:
+\begin{lemma}[Remainder Multiples] \label{lemma:1}
+    For given integers $a,b,n > 0$ with $a < b$, there exist exactly $(b - a) \div n$ integers $x \in [a,b)$ for every $0 \leq r < n$ such that $x \Mod n \equiv r$ whenever $n$ divides $b - a$.
+\end{lemma}
 
+%%% Might include later on - currently, there is no point in giving a overview over two sections
+\iffalse
 \subsection{Structure}
+We present and analyse existing techniques for uniformly sampling random integers in an interval $[0,n)$ in \hyperref[sec:2]{Section 2}.
+In \hyperref[sec:3]{Section 3}, we present the main contribution of this paper: a nearly-divisionless algorithm to sampler uniform random integers in an interval $[0,n)$.
+In \hyperref[sec:4]{Section 4}, we provide a short summary.
+\fi
 
 %%%%% Techniques
-\section{Existing Techniques}
+\section{Existing Techniques}\label{sec:2}
+In this section, we will review a number of different techniques used for sampling a uniform random integer in a given range $[0,n)$.
+Throughout this section, we use {\texttt rand()} to indicate the drawing of a uniform random $W$-bit number.
+
 \subsection{Biased}
+While our goal is to generate \emph{uniform} random numbers, there are also some techniques that generate biased - thus not uniform - random integers in a given range in exchange for much less complexity.
+
+\subsubsection{Modulo Reduction}
+The most common and under-egineered algorithm is the modulo reduction, namely
+\begin{center}
+    $\texttt{rand()} \Mod n$
+\end{center}
+Clearly, this method returns a random integer in the interval $[0,n)$.
+However, as seen in \hyperref[eq:1]{equation (1)}, for $n$ that do not perfectly divide $2^W$, we have a \emph{leftover} interval at the end that is biased.
+Thus, not all integers in $[0,n)$ are equally likely and this reduction does not produce uniform numbers.
+
+Furthermore, while this approach is probably the most intuitive and simplistic, it also illustrates why drawing numbers within a range is typically slow.
+We require a remainder operation which can be implemented through the division (\texttt{div}) instruction.
+For $32$-bit registers, this instruction has a latency of $26$ cycles and for $64$-bit registers, the latency lies within the range of $35$ to $88$ cycles~\cite{Instructions}.
+
+\subsubsection{Floating-Point Conversion}
+A common method to not rely on the costly \texttt{div} instruction is using a fixed-point floating-point conversion which consists of the following steps:
+\begin{enumerate}
+    \item Draw $x = \texttt{rand()}$ and convert $x$ to a floating-point number.
+    \item Multiply $x$ with $2^{-W}$ to generate a floating point number $y \in [0,1)$.
+    \item Multiply $y$ with $n$ and round it down to an integer before returning the result.
+\end{enumerate}
+Again, this method clearly generates a random integer in $[0,n)$ because $y$ is less than $1$ and we multiply with $n$ before rounding down.
+However, in the typical floating-point standard (IEEE 754) with $32$ bits, we have a mantisse of $24$ bits and hence can at best represent all values in $[0,2^{24})$ divided by $2^{24}$.
+Thus, we do not get the same accuracy as we would have with a $32$-bit number and hence can not generate all numbers in $[0,n)$ if $n > 2^{24}$.
+A solution would be to instead use double precision floating-point numbers that have a mantisse of $53$ bits and thus can represent all values in $[0,2^{53})$ divided by $2^{53}$.
+But this again is not enough if $W = 64$ where we require a mantisse of at least $64$ bits which is not accessible on all systems (long double).
+Furthermore, the conversion also comes at a price: the corresponding instructions have a latency of at least $6$ cycles on Skylake processors~\cite{Instructions}.
+
+\subsubsection{Multiply-and-Shift}
+We can adjust the prior floating-point conversion method to only use fixed-point arithmetic rather than floating-point arithmetic:
+\begin{center}
+    $\left(\texttt{rand()} \cdot n\right) \gg W$
+\end{center}
+where $\gg W$ is equivalent to dividing by $2^W$ as mentioned in \hyperref[sec:1.2]{section 1}.
+Clearly, this method and the floating-point method before work the same way, hence creating the same bias - but here, we multiply first before dividing by $2^W$~\cite{MultShift}.
+On the other hand, by multiplying first, we might get a number much bigger than $2^W$ and thus needing a new representation with more than $W$ bits.
+For $W = 32$, we simply switch to a $64$-bit number, but for $W = 64$, we have to switch to a $128$-bit number which is inherently slower than $32$-bit or $64$-bit numbers.
+Alternatively, for $W = 32$, a good compiler could compile this into a $32$-bit \texttt{mult} instruction which generates a $32$-bit number for the first $32$ bits and a $32$-bit number for the last $32$ bits.
 
 \subsection{Unbiased}
+While all prior approaches are very simplistic and always require only a handful of instructions, they also generate a bias which can vary depending on $n$.
+But since we want to generate a \emph{uniform} random number in $[0,n)$, we somehow need to get rid of such bias by incorporating a fail-safe: rejection sampling.
+
+\subsubsection{OpenBSD}
+We somehow want to get rid of the bias generated by the last interval in \hyperref[eq:1]{equation (1)}.
+The key idea is that this interval does not have to be on the right.
+Instead of rejecting all values generated by a random $x \geq 2^W - (2^W \Mod n)$, we could also reject all values generated by a random $x < 2^W \Mod n$ to achieve uniformity since the interval $[2^W \Mod n, 2^W)$ has size $2^W - (2^W \Mod n)$ which is divisible by $n$.
+Due to \hyperref[lemma:1]{Lemma 1}, for every $r \in [0,n)$, we have exactly $(b - a) \div n$ numbers in $[2^W \Mod n,2^W)$ with remainder $r$, thus having a uniform distribution over $[0,n)$ when computing the remainder.
+This yields \hyperref[alg:openbsd]{Algorithm 4} which stems from the C standard library in OpenBSD and macOS (\texttt{arc4random\_uniform}), and has been adopted by the Go language (\texttt{Int63n, Int31n}) with slight implementation differences~\cite{GoLang} as well as the GNU C++ standard library~\cite{GnuCpp}.
+
+%%% BEGIN OpenBSD
+\begin{algorithm}[!htb] \label{alg:openbsd}
+    \caption{The OpenBSD algorithm.}
+    \Require{source of uniformly-distributed random integers in $[0,2^W)$ given by \texttt{rand()}}
+    \Require{size of target interval given by $n$}
+    $t \leftarrow \left(2^W - n\right) \Mod n$\tcc*[r]{$ \left(2^W - n\right) \Mod n \equiv 2^W \Mod n$}
+    $x \leftarrow$ \texttt{rand()}\;
+    \While{$x < t$}{
+        $x \leftarrow$ \texttt{rand()}\;
+    }
+    \KwRet{$x \Mod n$}\;
+\end{algorithm}
+%%% END OpenBSD
+
+In comparison to the biased techniques, we now may have to sample more than one random integer in $[0,2^W)$ to generate an integer in $[0,n)$.
+Note that the number of random $W$-bit numbers drawn follows a geometric distribution with a success probability of $p = 1 - \frac{2^W \Mod n}{2^W}$.
+Thus on average, we need $\frac{1}{p}$ random $W$-bit numbers which is less than $2$ because $p > \frac{1}{2}$, irrespective of $n$.
+We also always require two remainder operations: one to compute the threshold that determines whether we reject a drawn random number and one to use the modulo reduction to generate the final number.
+
+\subsubsection{Java}
+Since we expect remainder operations to have high latency, always needing to compute two remainders can be quite costly.
+Hence, the Java language, in its \texttt{Random class}, uses a different approach.
+The key insight is that when we draw a random integer $x \in [0,2^W)$ and compute its remainder $r = x \Mod n$, we can map all integers in $[0,2^W)$ to a multiple of $n$ by computing $x - r$.
+Then, every number in the last interval in \hyperref[eq:1]{equation (1)} gets mapped to the last multiple $y$ of $n$ which is not greater than $2^W$.
+Note that this implies $y > 2^W - n$.
+Every other integer is mapped to a different (and thus smaller) multiple $z$ of $n$ for which $z \leq y - n$ and thus $z \leq 2^W - n$.
+Consequently, if we reject $x$ if $x - r > 2^W - n$ and otherwise return $r$, we get a unbiased generator.
+The details can be seen in \hyperref[alg:java]{Algorithm 5}.
+
+%%% BEGIN Java
+\begin{algorithm}[!htb] \label{alg:java}
+    \caption{The Java algorithm.}
+    \Require{source of uniformly-distributed random integers in $[0,2^W)$ given by \texttt{rand()}}
+    \Require{size of target interval given by $n$}
+    $x \leftarrow$ \texttt{rand()}\;
+    $r \leftarrow x \Mod n$\;
+    \While{$x - r > 2^W - n$}{
+        $x \leftarrow$ \texttt{rand()}\;
+        $r \leftarrow x \Mod n$\;
+    }
+    \KwRet{$r$}\;
+\end{algorithm}
+%%% END Java
+
+Again, the number of random draws of $W$-bit numbers follows a geometric distrbution with $p = 1 - \frac{2^W \Mod n}{2^W}$ and we expect to draw $\frac{1}{p} < 2$ random $W$-bit numbers before returning.
+However, this time, for every drawn random number, we also need to compute a remainder, hence we also expect $\frac{1}{p} < 2$ remainder operations.
+For small $n$, this is great because with high probability, we only need to compute one remainder operation in expectation as opposed to two.
+
+\subsubsection{Fast-Dice-Roller}
+For all prior unbiased algorithms, we need to use at least one remainder operation to debias the algorithm.
+Jérémie Lumbroso however devised an algorithm for sampling uniform random integers in $[0,n)$ for a given $n$ by iteratively building up the final number bit-by-bit~\cite{Flips}.
+In comparison to previous algorithms, we do not use the whole $W$ bits generated by the PRNG but instead only use one bit at a time building up the final number consisting of $\lceil\log_2n\rceil + 1$ bits.
+If we, at some point, overshoot and generate a number bigger than $n$ (but smaller than $2n$) we take a step back by subtracting $n$ and rebuilding the necessary bits.
+The algorithmic details can be seen in \hyperref[alg:flips]{Algorithm 6}.
+
+%%% BEGIN Flips
+\begin{algorithm}[!htb] \label{alg:flips}
+    \caption{The Fast-Dice-Roller algorithm.}
+    \Require{source of a uniformly-distributed random bit given by \texttt{flip()}}
+    \Require{size of target interval given by $n$}
+    $b \leftarrow 1$\;
+    $x \leftarrow 0$\;
+    \While{\emph{true}}{
+        $b \leftarrow 2 \cdot b$\;
+        $x \leftarrow 2 \cdot x + \texttt{flip()}$\;
+        \If{$b \geq n$}{
+            \uIf{$x < n$}{
+                \KwRet{$x$}\;
+            } \Else{
+                $b \leftarrow b - n$\;
+                $x \leftarrow x - n$\;
+            }
+        }
+    }
+\end{algorithm}
+%%% END Flips
+
+\begin{lemma}[Correctness of Fast-Dice-Roller~\cite{Flips}]
+    The Fast-Dice-Roller algorithms returns a uniformly distributed random integer in $[0,n)$ and terminates in expected time $\Theta(\log n)$.
+\end{lemma}
+\begin{proof}
+    Consider the following loop invariant: $x$ is uniformly distributed over $[0,b)$.
+    At the beginning, this is trivially true as $b$ is $1$ the only value $x$ can take (and takes) is $0$.
+    In line $4$ we double $b$, but also double $x$ in line $5$ before adding a parity bit which ensures uniformity in the enlarged new range.
+    Since we reach line $10$ and $11$ only when $x \geq n$, $x$ must be uniformly distributed in $[n,b)$ at this point.
+    Since we subtract $n$ from both $b$ and $x$, we simply shift the range from $[n,b)$ to $[0,b - n)$ maintaining uniformity.
+    As $x$ is always uniformly distributed in $[0,b)$ due to this loop invariant, when we return $x$, it must also be uniformly distributed over $[0,n)$.
+
+    For the proof of termination time, consider a less efficient variant of this algorithms that throws away all prior sampled bits when rejecting in line $9$-$11$.
+    Then, in each iteration, we have $\Theta(\log n)$ steps before the condition in line $6$ is \emph{true}.
+    Since we double $b$ in each iteration, we know that $b < 2n$ and hence, due to the previous loop invariant, we accept with probability $p = \nicefrac{n}{2^{\lfloor\log_2n\rfloor + 1}} > \nicefrac{1}{2}$, thus proving the lemma.
+\end{proof}
+Clearly, while this algorithm completely avoids any remainder or division operation, for most $n$, it is too slow.
+In comparison, every prior algorithm needs constant time or expected constant time to generate a uniform random number in $[0,n)$.
+Fast-Dice-Roller, on the other hand, needs expected logarithmic time for sampling which is catastrophical for bigger $n$ - a big price to pay for avoiding division (and multiplication if line $4$ and $5$ are implemented using a \textsc{BitShift}).
+
+\subsubsection{Bitmask}
+The main reason, Fast-Dice-Roller is highly impractical, is due to the fact that we build up our random number bit-by-bit instead of drawing all required bits at once.
+This motivates the following algorithm (see \hyperref[alg:bitmask]{Algorithm 7}) that was adopted by Apple in the macOS Sierra release when they made their own revision to the code of \texttt{arc4random\_uniform}~\cite{Apple}.
+
+%%% BEGIN Bitmask
+\begin{algorithm}[!htb] \label{alg:bitmask}
+    \caption{The Bitmask algorithm.}
+    \Require{source of uniformly-distributed random integers in $[0,2^W)$ given by \texttt{rand()}}
+    \Require{size of target interval given by $n$}
+    $z \leftarrow$ number of leading zeros in the binary representation of $n$\;
+    $m \leftarrow 1 \ll (W - z)$\tcc*[r]{$1 \ll (W - z) \equiv 2^{W - z}$}
+    $m \leftarrow m - 1$\;
+    $x \leftarrow \texttt{rand()} \And m$\;
+    \While{$x \geq n$}{
+        $x \leftarrow \texttt{rand()} \And m$\;
+    }
+    \KwRet{$x$}\;
+\end{algorithm}
+%%% END Bitmask
+
+The key idea is to first compute $k$ such that $k$ is the smallest value with $2^k \geq n$.
+We then can use a Bitmask $m = 2^k - 1$ to only take the necessary $k$ bits from our generated $W$-bit number and use rejection sampling to generate a uniform random number in $[0,n)$.
+This works in expected constant time since we accept with probability $p = \nicefrac{n}{2^k} > \nicefrac{1}{2}$.
+
+Bitmask completely avoids any type of division operation.
+Instead, we have to compute the number of leading zeros in line $1$.
+For C and C++, the GCC-Compiler provides the \texttt{\_\_builtin\_clz()} method to compute just that~\cite{GccCLZ}.
+In Rust, we can use the \texttt{leading\_zeros()} method of the standard library~\cite{RustCLZ}.
+These methods however can perform vastly different depending on the underlying architecture.
+For processors which have an inbuilt \texttt{clz} instruction such as ARM~\cite{ArmCLZ}, the compiler could translate this line into such an assembly instruction.
+In systems without direct support, one might to rely on more algorithmic implementations such as~\cite{ImplCLZ}.
 
 %%%%% Lemire
-\section{Nearly Divisionless Unbiased Sampling}
-
-%%%%% Experiments
-\section{Experiments}
-\subsection{Shuffling-Experiments}
-
-\subsection{In-Memory-Processing-Experiments}
+\section{Nearly Divisionless Unbiased Sampling}\label{sec:3}
 
 %%%%% Conclusion
-\section{Conclusion}
+\section{Conclusion}\label{sec:4}
 

### 2023-12-04 Paper/main.tex
section 3 + 4

@@
 \usepackage{hyperref}
+\usepackage{array}
+\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}} % Centered Multilines
+\renewcommand{\arraystretch}{1.5}
+\usepackage{booktabs} % Rules
+\usepackage{pifont} % checkmark / cross - thanks to `https://tex.stackexchange.com/questions/42619/xmark-that-complements-the-ams-checkmark`
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
 
@@
     A possible algorithm for this problem is Reservoir-Sampling~\cite{Reservoir}(see \hyperref[alg:gnm]{Algorithm 2}) which is a general technique that can sample $k$ elements without replacement from an input stream.
-    Another model is the BA model~\cite{BABook,BA} which iteratively builds the graph by adding nodes with edges to previously added nodes (see \hyperref[alg:ba]{Algorithm 3}). 
+    Another model is the BA model~\cite{BABook,BA} which iteratively builds the graph by adding nodes with edges to previously added nodes. %(see \hyperref[alg:ba]{Algorithm 3}). 
     Both models require sampling of an integer in an interval for each of its edges (or more).
@@
 
+%%% This block will not be included in the final version but is currently here in case the feedback says, it is useful and not over the top
+\iffalse
 %%% BEGIN BA-Sampling
@@
 %%% END BA-Sampling
+\fi
 
@@
 
-%%% Might include later on - currently, there is no point in giving a overview over two sections
-\iffalse
-\subsection{Structure}
-We present and analyse existing techniques for uniformly sampling random integers in an interval $[0,n)$ in \hyperref[sec:2]{Section 2}.
-In \hyperref[sec:3]{Section 3}, we present the main contribution of this paper: a nearly-divisionless algorithm to sampler uniform random integers in an interval $[0,n)$.
-In \hyperref[sec:4]{Section 4}, we provide a short summary.
-\fi
-
 %%%%% Techniques
@@
 
-\subsubsection{Modulo Reduction}
+\subsubsection{Modulo Reduction}\label{sec:2.1.1}
 The most common and under-egineered algorithm is the modulo reduction, namely
@@
 
-\subsubsection{Floating-Point Conversion}
+\subsubsection{Floating-Point Conversion}\label{sec:2.1.2}
 A common method to not rely on the costly \texttt{div} instruction is using a fixed-point floating-point conversion which consists of the following steps:
@@
 \end{enumerate}
+
 Again, this method clearly generates a random integer in $[0,n)$ because $y$ is less than $1$ and we multiply with $n$ before rounding down.
@@
 
-\subsubsection{Multiply-and-Shift}
+\subsubsection{Multiply-and-Shift}\label{sec:2.1.3}
 We can adjust the prior floating-point conversion method to only use fixed-point arithmetic rather than floating-point arithmetic:
@@
 Clearly, this method and the floating-point method before work the same way, hence creating the same bias - but here, we multiply first before dividing by $2^W$~\cite{MultShift}.
-On the other hand, by multiplying first, we might get a number much bigger than $2^W$ and thus needing a new representation with more than $W$ bits.
+On the other hand, by multiplying with $n$ first, we get a number in $[0,n \cdot 2^W)$ instead.
+If we now divide by $2^W$, we map all multiples of $n$ in $[0,2^W)$ to $0$, all multiples of $n$ in $[2^W, 2 \cdot 2^W)$ to $1$, and so on such that we map all multiples of $n$ in $[i \cdot 2^W, (i + 1) \cdot 2^W)$ (the $(i + 1)$\textsuperscript{th} $2^W$-interval) to $i$.
+
+Since we might get a number much bigger than $2^W$ and thus need a new representation with more than $W$ bits.
 For $W = 32$, we simply switch to a $64$-bit number, but for $W = 64$, we have to switch to a $128$-bit number which is inherently slower than $32$-bit or $64$-bit numbers.
@@
 
-\subsubsection{OpenBSD}
+\subsubsection{OpenBSD}\label{sec:2.2.1}
 We somehow want to get rid of the bias generated by the last interval in \hyperref[eq:1]{equation (1)}.
@@
 
-\subsubsection{Java}
+\subsubsection{Java}\label{sec:2.2.2}
 Since we expect remainder operations to have high latency, always needing to compute two remainders can be quite costly.
@@
 
-\subsubsection{Fast-Dice-Roller}
+\subsubsection{Fast-Dice-Roller}\label{sec:2.2.3}
 For all prior unbiased algorithms, we need to use at least one remainder operation to debias the algorithm.
@@
 
-\subsubsection{Bitmask}
+\subsubsection{Bitmask}\label{sec:2.2.4}
 The main reason, Fast-Dice-Roller is highly impractical, is due to the fact that we build up our random number bit-by-bit instead of drawing all required bits at once.
@@
 \section{Nearly Divisionless Unbiased Sampling}\label{sec:3}
+While completely avoiding division seems possible with the Fast-Dice-Roller and the Bitmask algorithm, it also comes at a cost: logarithmic time or a \texttt{clz} operation that might be even more expensive than a division on some systems.
+Hence, Lemire proposed an unbiased algorithm that uses the approach of the Multiply-and-Shift algorithm coupled with rejection sampling to remove bias.
+The algorithm (see \hyperref[alg:lemire]{Algorithm 8}) uses one divison operation at most and may also use no division operation at all.
+It is the main result of the underlying paper of this report~\cite{Lemire}.
+
+%%% BEGIN Lemire
+\begin{algorithm}[!htb] \label{alg:lemire}
+    \caption{The Lemire algorithm.}
+    \Require{source of uniformly-distributed random integers in $[0,2^W)$ given by \texttt{rand()}}
+    \Require{size of target interval given by $n$}
+    $x \leftarrow \texttt{rand()}$\;
+    $m \leftarrow x \cdot n$\;
+    $l \leftarrow m \Mod 2^W$\;
+    \If{$l < n$}{
+        $t \leftarrow \left(2^W - n\right) \Mod n$\tcc*[r]{$\left(2^W - n\right) \Mod n \equiv 2^W \Mod n$}
+        \While{$l < t$}{
+            $x \leftarrow \texttt{rand()}$\;
+        $m \leftarrow x \cdot n$\;
+        $l \leftarrow m \Mod 2^W$\;
+        }
+    }
+    \KwRet{$m \div 2^W$}\;
+\end{algorithm}
+%%% END Lemire
+
+\noindent Note that the remainder operation in line $3$ and $9$ can be computed by a logical \textsc{And} operation and the division operation in line $10$ by a Right-\textsc{BitShift} as mentioned in \hyperref[sec:1.2]{Section 1}
+Thus, the only \emph{real} division operation is the remainder operation in line $5$.
+
+Since we use the same mapping as in Multiply-and-Shift, we remind the reader of the $(i + 1)$\textsuperscript{th} $2^W$-interval introduced earlier: $[i \cdot 2^W, (i + 1) \cdot 2^W)$.
+By \hyperref[lemma:1]{Lemma 1}, there exist exactly $\lfloor\nicefrac{2^W}{n}\rfloor$ multiples of $n$ in the restricted $(i + 1)$\textsuperscript{th} $2^W$-interval $[i \cdot 2^W + (2^W \Mod n), (i + 1) \cdot 2^W)$ as $n$ perfectly divides $(i + 1) \cdot 2^W - (i \cdot 2^W + (2^W \Mod n)) = 2^W - (2^W \Mod n)$.
+Thus, we can remove bias if we reject all multiples of $n$ in the interval $[i \cdot 2^W, i \cdot 2^W + (2^W \Mod n))$ for every $i$.
+
+The above algorithm does just that in lines $5$ and $6$.
+But it also makes use of the following observation: since $t = 2^W \Mod n$, we have $t < n$ and thus if we have $l \geq n$, we also know that $l \geq t$.
+In this case we know that we do not have to reject the current value without computing $t$ in the first place.
+Thus, we can completely skip the division operation with probability $p = \frac{2^W - n}{2^W} = 1 - \frac{n}{2^W}$.
 
@@
 \section{Conclusion}\label{sec:4}
+We have seen several different algorithms that can be used to sample a random integer in an interval $[0,n)$.
+The results are summarized in \hyperref[tab:1]{Table 1}.
+As we introduce rejection sampling for unbiased algorithms, runtimes for those are taken as expected constant or logarithmic time, while runtimes for biased algorithms are constant overall.
+
+%%% BEGIN Results
+\begin{table}[!htb]
+    \centering
+    \begin{tabular}{c|P{25mm}P{25mm}P{15mm}P{15mm}}
+        \toprule
+        & expected number of integer division operations & maximal number of integer division operations & Unbiased? & (Expected) Runtime \\
+        \midrule
+        \hyperref[sec:2.1.1]{Modulo Reduction} & $1$ & $1$ & \xmark & $\Theta(1)$ \\
+        \hyperref[sec:2.1.2]{Floating-Point Conversion} & $0$ & $0$ & \xmark & $\Theta(1)$ \\
+        \hyperref[sec:2.1.3]{Multiply-and-Shift} & $0$ & $0$ & \xmark & $\Theta(1)$ \\
+        \hyperref[sec:2.2.1]{OpenBSD} & $2$ & $2$ & \cmark & $\Theta(1)$ \\
+        \hyperref[sec:2.2.2]{Java} & $\frac{2^W}{2^W - (2^W \Mod n)}$ & $\infty$ & \cmark & $\Theta(1)$ \\
+        \hyperref[sec:2.2.3]{Fast-Dice-Roller} & $0$ & $0$ & \cmark & $\Theta(\log n)$ \\
+        \hyperref[sec:2.2.4]{Bitmask} & $0$ & $0$ & \cmark & $\Theta(1)$ \\
+        \hyperref[sec:3]{Lemire} & $\frac{n}{2^W}$ & $1$ & \cmark & $\Theta(1)$ \\
+        \bottomrule
+    \end{tabular}
+    \caption{
+        Overview over different sampling algorithms: expected/maximal number of division operations, unbiased or biased, (expected) runtime
+    }
+    \label{tab:1}
+\end{table}
+%%% END Results
+
+While the bias is usually very small and negligible in most use-cases, if we sample millions of times, it can easily become apparent and strongly distort the final dataset depending on the algorithm.
+Hence, it is important to generate unbiased random integers at the cost of a little execution time.
+Common software libraries seem to rely on algorithms that always require at least one division, if not two or more which is very inefficient on most processors.
+While Bitmask has in theory zero division operations, the required \texttt{clz} operation can be as inefficient as the \texttt{div} instruction.
+Hence, the algorithm introduced by Lemire is a perfect fit for modern x64 processors as for small $n$, we require no division (or equally bad) operation at all and also bound the maximum number of divisions to at most one.
+
+For different technologies however, support for the computation of the full multiplication might be lacking such as in Graphics Processing Units (GPUs)~\cite{GPU} or DRAM Processing Units (DPUs) used for In-Memory-Processing in the \textsc{Upmem} architecture~\cite{Upmem}.
+In the \textsc{Upmem-Pim} toolchain, for example, an \texttt{clz} operation seems to greatly outshine a \texttt{div} operation~\cite{UpmemRng}.
 

### 2023-12-04 Paper/main.tex
small changes

@@
 \subsection{Unbiased}
-While all prior approaches are very simplistic and always require only a handful of instructions, they also generate a bias which can vary depending on $n$.
+While all prior approaches are quite simplistic and always require only a handful of instructions, they also generate a bias which can vary depending on $n$.
 But since we want to generate a \emph{uniform} random number in $[0,n)$, we somehow need to get rid of such bias by incorporating a fail-safe: rejection sampling.
@@
 The key idea is that this interval does not have to be on the right.
-Instead of rejecting all values generated by a random $x \geq 2^W - (2^W \Mod n)$, we could also reject all values generated by a random $x < 2^W \Mod n$ to achieve uniformity since the interval $[2^W \Mod n, 2^W)$ has size $2^W - (2^W \Mod n)$ which is divisible by $n$.
+Instead of rejecting all values generated by a random $x \geq 2^W - (2^W \Mod n)$, we could also reject all values generated by a random $x < 2^W \Mod n$ for uniformity since the interval $[2^W \Mod n, 2^W)$ has size $2^W - (2^W \Mod n)$ which is divisible by $n$.
 Due to \hyperref[lemma:1]{Lemma 1}, for every $r \in [0,n)$, we have exactly $(b - a) \div n$ numbers in $[2^W \Mod n,2^W)$ with remainder $r$, thus having a uniform distribution over $[0,n)$ when computing the remainder.
@@
 
-While the bias is usually very small and negligible in most use-cases, if we sample millions of times, it can easily become apparent and strongly distort the final dataset depending on the algorithm.
+While the bias is usually insignificant and negligible in most use-cases, if we sample millions of times, it can easily become apparent and strongly distort the final dataset depending on the algorithm.
 Hence, it is important to generate unbiased random integers at the cost of a little execution time.
-Common software libraries seem to rely on algorithms that always require at least one division, if not two or more which is very inefficient on most processors.
+Common software libraries seem to rely on algorithms that always require at least one division, if not two or more which is inefficient on most processors.
 While Bitmask has in theory zero division operations, the required \texttt{clz} operation can be as inefficient as the \texttt{div} instruction.
@@
 
-For different technologies however, support for the computation of the full multiplication might be lacking such as in Graphics Processing Units (GPUs)~\cite{GPU} or DRAM Processing Units (DPUs) used for In-Memory-Processing in the \textsc{Upmem} architecture~\cite{Upmem}.
+For different technologies however, support for the computation of the full multiplication might be lacking such as in Graphics Processing Units (GPUs)~\cite{GPU} or DRAM Processing Units (DPUs) used for In-Memory-Processing in the \textsc{Upmem} architecture~\cite{Upmem, UpmemSDK}.
 In the \textsc{Upmem-Pim} toolchain, for example, an \texttt{clz} operation seems to greatly outshine a \texttt{div} operation~\cite{UpmemRng}.

### 2023-12-04 Paper/main.tex
fix algo numbering!

@@
 Due to \hyperref[lemma:1]{Lemma 1}, for every $r \in [0,n)$, we have exactly $(b - a) \div n$ numbers in $[2^W \Mod n,2^W)$ with remainder $r$, thus having a uniform distribution over $[0,n)$ when computing the remainder.
-This yields \hyperref[alg:openbsd]{Algorithm 4} which stems from the C standard library in OpenBSD and macOS (\texttt{arc4random\_uniform}), and has been adopted by the Go language (\texttt{Int63n, Int31n}) with slight implementation differences~\cite{GoLang} as well as the GNU C++ standard library~\cite{GnuCpp}.
+This yields \hyperref[alg:openbsd]{Algorithm 3} which stems from the C standard library in OpenBSD and macOS (\texttt{arc4random\_uniform}), and has been adopted by the Go language (\texttt{Int63n, Int31n}) with slight implementation differences~\cite{GoLang} as well as the GNU C++ standard library~\cite{GnuCpp}.
 
@@
 Consequently, if we reject $x$ if $x - r > 2^W - n$ and otherwise return $r$, we get a unbiased generator.
-The details can be seen in \hyperref[alg:java]{Algorithm 5}.
+The details can be seen in \hyperref[alg:java]{Algorithm 4}.
 
@@
 If we, at some point, overshoot and generate a number bigger than $n$ (but smaller than $2n$) we take a step back by subtracting $n$ and rebuilding the necessary bits.
-The algorithmic details can be seen in \hyperref[alg:flips]{Algorithm 6}.
+The algorithmic details can be seen in \hyperref[alg:flips]{Algorithm 5}.
 
@@
 The main reason, Fast-Dice-Roller is highly impractical, is due to the fact that we build up our random number bit-by-bit instead of drawing all required bits at once.
-This motivates the following algorithm (see \hyperref[alg:bitmask]{Algorithm 7}) that was adopted by Apple in the macOS Sierra release when they made their own revision to the code of \texttt{arc4random\_uniform}~\cite{Apple}.
+This motivates the following algorithm (see \hyperref[alg:bitmask]{Algorithm 6}) that was adopted by Apple in the macOS Sierra release when they made their own revision to the code of \texttt{arc4random\_uniform}~\cite{Apple}.
 
@@
 Hence, Lemire proposed an unbiased algorithm that uses the approach of the Multiply-and-Shift algorithm coupled with rejection sampling to remove bias.
-The algorithm (see \hyperref[alg:lemire]{Algorithm 8}) uses one divison operation at most and may also use no division operation at all.
+The algorithm (see \hyperref[alg:lemire]{Algorithm 7}) uses one divison operation at most and may also use no division operation at all.
 It is the main result of the underlying paper of this report~\cite{Lemire}.

### 2023-12-04 Paper/main.tex
Abstract

@@
 \begin{abstract}
-    Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Praesent convallis orci arcu, eu mollis dolor.
-Aliquam eleifend suscipit lacinia.
-Maecenas quam mi, porta ut lacinia sed, convallis ac dui.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse potenti.
+    In randomized algorithms and other probabilistic practices we often need to generate uniform random integers in a given interval.
+    This is essential for algorithms like the Fisher-Yates random shuffle or simply drawing an element from a list uniformly at random.
+    Pseudo-random numbers however are often generated with a fixed number of bits and thus need to be mapped to the desired interval without introducing bias.
+    Common methods often include division operations that are relatively expensive on most processors.
+    Therefore, Lemire introduced an algorithm that can sample uniform random integers in a given range $[0,n)$ with zero to one division operations at most~\cite{Lemire}.
+
+    In this seminar report on the paper by Lemire, we present a variety of algorithms that generate random integers in a given range and compare them on runtime, number of division operations, and statistical bias.
 \end{abstract}

### 2024-01-09 Paper/main.tex
incorporate feedback

--- 2023-12-04 Paper/main.tex
@@
 \begin{abstract}
-    Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Praesent convallis orci arcu, eu mollis dolor.
-Aliquam eleifend suscipit lacinia.
-Maecenas quam mi, porta ut lacinia sed, convallis ac dui.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse potenti.
+    In randomized algorithms and other probabilistic methods we often need to generate uniform random integers in a given interval.
+    This is essential for algorithms like the Fisher-Yates random shuffle or simply drawing an element from an array uniformly at random.
+    Pseudo-random numbers however are often generated as words with a fixed number of bits and thus need to be mapped to the desired interval without introducing bias.
+    Common methods often include division operations that are relatively expensive on most processors.
+    Therefore, Lemire introduced an algorithm that can sample uniform random integers in a given range $[0,n)$ with zero to one division operations at most~\cite{Lemire}.
+
+    In this report, we present a variety of algorithms that generate random integers in a given range and compare them on runtime, number of division operations, and statistical bias.
 \end{abstract}
@@
 \subsection{Motivation}
-The study of randomized algorithms goes as far back as the 1910s~\cite{MathComp}.
-While initially small, this study is now scientifically pursued in almost all parts of computer science.
+The study of randomized algorithms goes as far back as the 1910s~\cite{MathComp} and is now scientifically pursued in almost all parts of computer science.
 Not only in theory, but also in practice, randomized algorithms gain even more importance today to not only solve specific algorithmic problems but also to create large, scalable, and unbiased data sets that can be used in a variety of fields.
 \\
-The core of such algorithms is most often the pseudo-random number generator (PRNG).
-There are many efficient PRNGs such as XorShift~\cite{XS}, MersenneTwister~\cite{MT}, linear congruential generators~\cite{LinConGen, MINSTD, LM}, and younger generators such as the PCG-Family~\cite{PCG} and ChaCha~\cite{Cha}.
+The core of such algorithms is often the pseudo-random number generator (PRNG).
+There are many efficient PRNGs such as XorShift~\cite{XS}, MersenneTwister~\cite{MT}, linear congruential generators~\cite{LinConGen, MINSTD, LM}, and younger generators such as the PCG-family~\cite{PCG} and ChaCha~\cite{Cha}.
 These generators produce $32$-bit or $64$-bit words which we can interpret as unsigned integers in $[0,2^{32})$ and $[0,2^{64})$ respectively.
@@
 \begin{itemize}
-    \item The Fisher-Yates random shuffle~\cite{FY} (see \hyperref[alg:fy]{Algorithm 1}) uniformly permutes an array of $n$ elements in time $\Theta(n)$. 
+    \item The Fisher-Yates random shuffle~\cite{FY} (see \hyperref[alg:fy]{Algorithm 1}) uniformly permutes an array of $n$ elements in time $\Theta(n)$ if we can sample in constant time. 
     We have to draw a random integer in an interval for every shuffled element.
@@
     For example, the $\mathcal{G}(n,m)$ model~\cite{GNM} uniformly samples a (directed) graph with $n$ nodes and $m$ edges. 
-    A possible algorithm for this problem is Reservoir-Sampling~\cite{Reservoir}(see \hyperref[alg:gnm]{Algorithm 2}) which is a general technique that can sample $k$ elements without replacement from an input stream.
+    A possible algorithm for this problem is Reservoir-Sampling (see \hyperref[alg:gnm]{Algorithm 2})~\cite{Reservoir} which is a general technique that can sample $k$ elements without replacement from an input stream.
     Another model is the BA model~\cite{BABook,BA} which iteratively builds the graph by adding nodes with edges to previously added nodes. %(see \hyperref[alg:ba]{Algorithm 3}). 
@@
     \item An Alias-Table~\cite{Alias} can be used to simulate static discrete distributions in practice. 
-    After creating and initializing the table, we can sample in (expected) time $\Theta(1)$ by drawing a uniform random row and offset in each round. 
+    After creating and initializing the table, we can sample in time $\Theta(1)$ by drawing a uniform random row and offset in each round. 
     Hence, we need to sample an integer uniformly at random in $[0,n)$ where $n$ is the number of rows in the table.
@@
     \For{$i = n - 1,\ldots,1$}{
-        $j \leftarrow$ random integer in $[0,i]$\;
+        $j \leftarrow$ \texttt{random integer in $[0,i]$}\;
         swap $A[i]$ and $A[j]$\;
@@
     \For{$i = m,\ldots,n^2$}{
-        $j \leftarrow$ random integer in $[0,i]$\;
+        $j \leftarrow$ \texttt{random integer in $[0,i]$}\;
         \If{$j < m$}{
@@
 \subsection{Preliminaries}\label{sec:1.2}
-We denote by $x \div y$ the integer division operation, namely $\lfloor\nicefrac{x}{y}\rfloor$, and the remainder operation by $x \Mod y \equiv x - (x \div y)y$. 
-Divisions by a power of two can be efficiently implemented by Bit-\textsc{RightShift} denoted by $\gg$: $x \div 2^W \equiv x \gg W$.
+We denote by $x \div y \coloneqq \lfloor\nicefrac{x}{y}\rfloor$ the integer division operation, and the remainder operation by $x \Mod y \coloneqq x - (x \div y)y$. 
+Using a Bit-\textsc{RightShift}, denoted by $\gg$, we can efficiently implement divisions by a power of two: $x \div 2^W \coloneqq x \gg W$.
 Similarly, we denote a Bit-\textsc{LeftShift} by $\ll$.
-Additionally, we can efficiently compute the remainder of a division py a power of two by a logical \textsc{And}-operation, denoted by $\&$: $x \Mod 2^W \equiv x \And (2^W - 1)$.
-Typically, PRNGs generate uniform random numbers in an interval $[0,2^W)$ where $W \in \{32,64\}$, namely a uniform random $W$-bit number.
+Additionally, with a bitwise \textsc{And}-operation, denoted by $\&$, we can efficiently compute the remainder of a division by a power of two: $x \Mod 2^W \coloneqq x \And (2^W - 1)$.
+In practice, PRNGs generate uniform random numbers in an interval $[0,2^W)$ where $W \in \{32,64\}$, namely a uniform random $W$-bit number.
 
-Since we can map every integer in an interval $[a,b)$ to the interval $[0,b - a)$ by subtracting $a$, we want to generate a uniform random integer in an interval $[0,n)$ for a given $n$.
+Since we can map every integer in an interval $[a,b)$ to the interval $[0,b - a)$ by subtracting $a$, we focus on generating a uniform random integer in an interval $[0,n)$ for a given $n$.
 If we map all integers $x$ in $[0,2^W)$ with the function $x \Mod n$, we get the following sequence:
@@
 }%
-This motivates the following lemma:
+Clearly, when $n$ does not divide $2^W$, integers in $[0,2^W \Mod n)$ have exactly one more occurence than integers in $[2^W \Mod n, n)$.
+When $n$ however does divide $2^W$, every integer in $[0,n)$ has exactly the same number of occurences.
+Generalizing this notion yields the following lemma:
 \begin{lemma}[Remainder Multiples] \label{lemma:1}
@@
 In this section, we will review a number of different techniques used for sampling a uniform random integer in a given range $[0,n)$.
-Throughout this section, we use {\texttt rand()} to indicate the drawing of a uniform random $W$-bit number.
+Throughout this section, we use \texttt{rand()} to indicate the drawing of a uniform random $W$-bit number.
 
 \subsection{Biased}
-While our goal is to generate \emph{uniform} random numbers, there are also some techniques that generate biased - thus not uniform - random integers in a given range in exchange for much less complexity.
+While our goal is to generate \emph{uniform} random numbers, there are also some techniques that generate biased --- thus not uniform --- random integers in a given range in exchange for much less complexity.
 
@@
 Clearly, this method returns a random integer in the interval $[0,n)$.
-However, as seen in \hyperref[eq:1]{equation (1)}, for $n$ that do not perfectly divide $2^W$, we have a \emph{leftover} interval at the end that is biased.
+However, as seen in \hyperref[eq:1]{equation (1)}, if $n$ does not perfectly divide $2^W$, we have a \emph{leftover} interval at the end that introduces bias as not all integers in $[0,n)$ are a part of this interval.
 Thus, not all integers in $[0,n)$ are equally likely and this reduction does not produce uniform numbers.
@@
 Furthermore, while this approach is probably the most intuitive and simplistic, it also illustrates why drawing numbers within a range is typically slow.
-We require a remainder operation which can be implemented through the division (\texttt{div}) instruction.
+We require a remainder operation which can be implemented through the division (\texttt{div}) instruction on x64 processors.
 For $32$-bit registers, this instruction has a latency of $26$ cycles and for $64$-bit registers, the latency lies within the range of $35$ to $88$ cycles~\cite{Instructions}.
@@
 Again, this method clearly generates a random integer in $[0,n)$ because $y$ is less than $1$ and we multiply with $n$ before rounding down.
-However, in the typical floating-point standard (IEEE 754) with $32$ bits, we have a mantisse of $24$ bits and hence can at best represent all values in $[0,2^{24})$ divided by $2^{24}$.
-Thus, we do not get the same accuracy as we would have with a $32$-bit number and hence can not generate all numbers in $[0,n)$ if $n > 2^{24}$.
+However, in the commonly implemented floating-point standard (IEEE 754) with $32$ bits, we have a mantisse of $24$ bits and hence can at best represent all values in $[0,2^{24})$ divided by $2^{24}$.
+Thus, we can not represent all numbers in $[0,2^{32})$ divided by $2^{32}$ and have less precision than a $32$-bit number.
+Hence can not generate all numbers in $[0,n)$ if $n > 2^{24}$.
 A solution would be to instead use double precision floating-point numbers that have a mantisse of $53$ bits and thus can represent all values in $[0,2^{53})$ divided by $2^{53}$.
@@
 where $\gg W$ is equivalent to dividing by $2^W$ as mentioned in \hyperref[sec:1.2]{section 1}.
-Clearly, this method and the floating-point method before work the same way, hence creating the same bias - but here, we multiply first before dividing by $2^W$~\cite{MultShift}.
+Clearly, this method and the floating-point method before work the same way, hence creating the same bias --- but here, we multiply first before dividing by $2^W$~\cite{MultShift}.
 On the other hand, by multiplying with $n$ first, we get a number in $[0,n \cdot 2^W)$ instead.
-If we now divide by $2^W$, we map all multiples of $n$ in $[0,2^W)$ to $0$, all multiples of $n$ in $[2^W, 2 \cdot 2^W)$ to $1$, and so on such that we map all multiples of $n$ in $[i \cdot 2^W, (i + 1) \cdot 2^W)$ (the $(i + 1)$\textsuperscript{th} $2^W$-interval) to $i$.
+If we now divide by $2^W$, we map all multiples of $n$ in $[0,2^W)$ to $0$, all multiples of $n$ in $[2^W, 2 \cdot 2^W)$ to $1$, and so on, that is, we map all multiples of $n$ in $[i \cdot 2^W, (i + 1) \cdot 2^W)$ (the $(i + 1)$\textsuperscript{th} $2^W$-interval) to $i$.
 
-Since we might get a number much bigger than $2^W$ and thus need a new representation with more than $W$ bits.
-For $W = 32$, we simply switch to a $64$-bit number, but for $W = 64$, we have to switch to a $128$-bit number which is inherently slower than $32$-bit or $64$-bit numbers.
+Since multiplying a random number in $[0,2^W)$ by $n$ might yield a number bigger than $2^W$ we need a new representation with more than $W$ bits.
+For $W = 32$, we simply extend to a $64$-bit number, but for $W = 64$, we have to use a $128$-bit number which does not fit into a single register on any machine. 
+This makes $128$-bit numbers inherently slower than $32$-bit or $64$-bit numbers.
 Alternatively, for $W = 32$, a good compiler could compile this into a $32$-bit \texttt{mult} instruction which generates a $32$-bit number for the first $32$ bits and a $32$-bit number for the last $32$ bits.
@@
 While all prior approaches are quite simplistic and always require only a handful of instructions, they also generate a bias which can vary depending on $n$.
-But since we want to generate a \emph{uniform} random number in $[0,n)$, we somehow need to get rid of such bias by incorporating a fail-safe: rejection sampling.
+But since we want to generate a \emph{uniform} random number in $[0,n)$, we somehow need to get rid of such bias by introducing rejection sampling.
 
@@
 The key idea is that this interval does not have to be on the right.
-Instead of rejecting all values generated by a random $x \geq 2^W - (2^W \Mod n)$, we could also reject all values generated by a random $x < 2^W \Mod n$ for uniformity since the interval $[2^W \Mod n, 2^W)$ has size $2^W - (2^W \Mod n)$ which is divisible by $n$.
+Instead of rejecting all values generated by a random $x \geq 2^W - (2^W \Mod n)$, we could also reject all values generated by a random $x < 2^W \Mod n$ for uniformity since the interval $[2^W \Mod n, 2^W)$ is continuous and has size $2^W - (2^W \Mod n)$ which is divisible by $n$.
 Due to \hyperref[lemma:1]{Lemma 1}, for every $r \in [0,n)$, we have exactly $(b - a) \div n$ numbers in $[2^W \Mod n,2^W)$ with remainder $r$, thus having a uniform distribution over $[0,n)$ when computing the remainder.
@@
 \subsubsection{Fast-Dice-Roller}\label{sec:2.2.3}
-For all prior unbiased algorithms, we need to use at least one remainder operation to debias the algorithm.
+For all prior unbiased algorithms, we need to use at least one remainder operation to get rid of the bias in the algorithm.
 Jérémie Lumbroso however devised an algorithm for sampling uniform random integers in $[0,n)$ for a given $n$ by iteratively building up the final number bit-by-bit~\cite{Flips}.
@@
     \While{\emph{true}}{
-        $b \leftarrow 2 \cdot b$\;
-        $x \leftarrow 2 \cdot x + \texttt{flip()}$\;
+        $b \leftarrow b \ll 1$\;
+        $x \leftarrow (x \ll 1) \mid \texttt{flip()}$\;
         \If{$b \geq n$}{
@@
     At the beginning, this is trivially true as $b$ is $1$ the only value $x$ can take (and takes) is $0$.
-    In line $4$ we double $b$, but also double $x$ in line $5$ before adding a parity bit which ensures uniformity in the enlarged new range.
+    In line $4$ we double $b$, but also double $x$ in line $5$ before randomly setting the final bit which ensures uniformity in the enlarged new range.
     Since we reach line $10$ and $11$ only when $x \geq n$, $x$ must be uniformly distributed in $[n,b)$ at this point.
@@
 In comparison, every prior algorithm needs constant time or expected constant time to generate a uniform random number in $[0,n)$.
-Fast-Dice-Roller, on the other hand, needs expected logarithmic time for sampling which is catastrophical for bigger $n$ - a big price to pay for avoiding division (and multiplication if line $4$ and $5$ are implemented using a \textsc{BitShift}).
+Fast-Dice-Roller, on the other hand, needs expected logarithmic time for sampling which is catastrophical for bigger $n$.
 
 \subsubsection{Bitmask}\label{sec:2.2.4}
-The main reason, Fast-Dice-Roller is highly impractical, is due to the fact that we build up our random number bit-by-bit instead of drawing all required bits at once.
+Another method for completely avoiding division operations comes from the following observation:
+for every number less than $n$, only the first $\lceil\log_2(n)\rceil$ bits can be non-zero in an unsigned representation.
 This motivates the following algorithm (see \hyperref[alg:bitmask]{Algorithm 6}) that was adopted by Apple in the macOS Sierra release when they made their own revision to the code of \texttt{arc4random\_uniform}~\cite{Apple}.
@@
 For processors which have an inbuilt \texttt{clz} instruction such as ARM~\cite{ArmCLZ}, the compiler could translate this line into such an assembly instruction.
-In systems without direct support, one might to rely on more algorithmic implementations such as~\cite{ImplCLZ}.
+In systems without direct support, one might to rely on more algorithmic implementations such as~\cite{ImplCLZ, HackersDelight}.
+
+For small $n$, it might be inefficient to only use $\lceil\log_2(n)\rceil$ bits from each generated random $W$-bit number.
+Hence, using a buffer to store all random bits can speed up this algorithm as done in \texttt{arc4random\_uniform}~\cite{Apple}.
+Wu et al. provided a buffered version of this algorithm in~\cite{BitParallelFDR} which can also be interpreted as a bit-parallel Fast-Dice-Roller.
 
@@
 It is the main result of the underlying paper of this report~\cite{Lemire}. 
+\vspace*{0.5cm}
 
@@
     $x \leftarrow \texttt{rand()}$\;
-    $m \leftarrow x \cdot n$\;
+    $m \leftarrow x \cdot n$\tcc*[r]{$m$ needs $2 \cdot W$ bits for representation}
     $l \leftarrow m \Mod 2^W$\;
@@
 
-\noindent Note that the remainder operation in line $3$ and $9$ can be computed by a logical \textsc{And} operation and the division operation in line $10$ by a Right-\textsc{BitShift} as mentioned in \hyperref[sec:1.2]{Section 1}
+\noindent Note that the remainder operation in line $3$ and $9$ can be computed by a bitwise \textsc{And} operation and the division operation in line $10$ by a Right-\textsc{BitShift} as mentioned in \hyperref[sec:1.2]{Section 1}
 Thus, the only \emph{real} division operation is the remainder operation in line $5$.

### 2024-01-09 Paper/main.tex
remove algorithm

@@
 %%% END G(n,m) Sampling
-
-%%% This block will not be included in the final version but is currently here in case the feedback says, it is useful and not over the top
-\iffalse
-%%% BEGIN BA-Sampling
-\begin{algorithm}[!htb] \label{alg:ba}
-    \caption{Preferential-Attachment-Generator \cite{BASampling}: sample BA-graphs in linear time}
-    \Require{source of uniformly-distributed random integers in $[0,i]$ for parameter $i$}
-    \Require{inital edge-list $E_0$ with $n_0$ nodes and $m_0 \coloneqq |E_0|$ edges, number of additional nodes $N$, neighbor-parameter $\nu$}
-    $M \leftarrow$ array of size $2\left(m_0 + \nu N\right)$\;
-    \For{$i = 0,\ldots,m_0 - 1$}{
-        $\left(M[2i], M[2i + 1]\right) \leftarrow E_0[i]$\;
-    }
-    \For{$i = 0,\ldots,N - 1$}{
-        \For{$j = 1,\ldots,\nu$}{
-            $k \leftarrow$ random integer in $[0,2(m_0 + i\nu) - 1]$\;
-            $M[2(m_0 + i\nu)] \leftarrow M[k]$\;
-            $M[2(m_0 + i\nu) + 1] \leftarrow n_0 + i$\;
-        }
-    }
-    $E \leftarrow$ array (of edges) of size $\left(m_0 + \nu N\right)$\;
-    \For{$i = 0,\ldots,m_0 + \nu N - 1$}{
-        $E[i] \leftarrow (M[2i], M[2i + 1])$\;
-    }
-    \KwRet{$E$}\;
-\end{algorithm}
-%%% END BA-Sampling
-\fi
 

### 2024-01-09 Paper/main.tex
...

@@
 We require a remainder operation which can be implemented through the division (\texttt{div}) instruction on x64 processors.
-For $32$-bit registers, this instruction has a latency of $26$ cycles and for $64$-bit registers, the latency lies within the range of $35$ to $88$ cycles~\cite{Instructions}.
+On Intel Skylake processors, this instruction has a latency of $26$ cycles for $32$-bit registers and for $64$-bit registers, the latency lies within the range of $35$ to $88$ cycles.
+Usually, a \texttt{div} instruction is one to two orders of magnitude slower than an \texttt{add} instruction~\cite{Instructions}.
 
@@
 Hence, using a buffer to store all random bits can speed up this algorithm as done in \texttt{arc4random\_uniform}~\cite{Apple}.
-Wu et al. provided a buffered version of this algorithm in~\cite{BitParallelFDR} which can also be interpreted as a bit-parallel Fast-Dice-Roller.
+Wu et al. provided a buffered version of this algorithm in which can also be interpreted as a bit-parallel Fast-Dice-Roller~\cite{BitParallelFDR}.
 

### 2024-01-16 Paper/main.tex
finalize report

@@
 \relatedversion{} 
-\supplement{Sources of implementations and experiments can be accessed at \url{https://github.com/lukasgeis/FRIGiaI-Review}}
+\supplement{Sources can be accessed at \url{https://github.com/lukasgeis/FRIGiaI-Review}}
 \acknowledgements{}
@@
 \begin{abstract}
-    In randomized algorithms and other probabilistic methods we often need to generate uniform random integers in a given interval.
+    In randomized algorithms and other probabilistic methods, we often need to generate uniform random integers in a given interval.
     This is essential for algorithms like the Fisher-Yates random shuffle or simply drawing an element from an array uniformly at random.
@@
 However, for most applications we do not want to have such a large range of possible random values but instead want to confine the range to a given interval $[a,b]$.
-This can not only be used to uniformly sample an element from an array but also for more complex algorithms
+This can not only be used to uniformly sample an element from an array but also for more complex algorithms:
 \begin{itemize}
-    \item The Fisher-Yates random shuffle~\cite{FY} (see \hyperref[alg:fy]{Algorithm 1}) uniformly permutes an array of $n$ elements in time $\Theta(n)$ if we can sample in constant time. 
+    \item The Fisher-Yates random shuffle (see \hyperref[alg:fy]{Algorithm 1})~\cite{FY} uniformly permutes an array of $n$ elements in time $\Theta(n)$ if we can sample in constant time. 
     We have to draw a random integer in an interval for every shuffled element.
@@
 Using a Bit-\textsc{RightShift}, denoted by $\gg$, we can efficiently implement divisions by a power of two: $x \div 2^W \coloneqq x \gg W$.
-Similarly, we denote a Bit-\textsc{LeftShift} by $\ll$.
+Similarly, we denote a Bit-\textsc{LeftShift} by $\ll$ which can be used for efficient multiplication by a power of two.
 Additionally, with a bitwise \textsc{And}-operation, denoted by $\&$, we can efficiently compute the remainder of a division by a power of two: $x \Mod 2^W \coloneqq x \And (2^W - 1)$.
@@
 Clearly, this method returns a random integer in the interval $[0,n)$.
-However, as seen in \hyperref[eq:1]{equation (1)}, if $n$ does not perfectly divide $2^W$, we have a \emph{leftover} interval at the end that introduces bias as not all integers in $[0,n)$ are a part of this interval.
+However, as seen in \hyperref[eq:1]{equation (1)}, if $n$ does not divide $2^W$, we have a \emph{leftover} interval at the end that introduces bias as not all integers in $[0,n)$ are a part of this interval.
 Thus, not all integers in $[0,n)$ are equally likely and this reduction does not produce uniform numbers.
@@
 
-Again, this method clearly generates a random integer in $[0,n)$ because $y$ is less than $1$ and we multiply with $n$ before rounding down.
+Again, this method clearly generates a random integer in $[0,n)$ because $y$ is at most $1.0$ and we multiply with $n$ before rounding down (floating-point-errors are negligible because they get corrected in (3) by rounding down).
 However, in the commonly implemented floating-point standard (IEEE 754) with $32$ bits, we have a mantisse of $24$ bits and hence can at best represent all values in $[0,2^{24})$ divided by $2^{24}$.
@@
 Instead of rejecting all values generated by a random $x \geq 2^W - (2^W \Mod n)$, we could also reject all values generated by a random $x < 2^W \Mod n$ for uniformity since the interval $[2^W \Mod n, 2^W)$ is continuous and has size $2^W - (2^W \Mod n)$ which is divisible by $n$.
-Due to \hyperref[lemma:1]{Lemma 1}, for every $r \in [0,n)$, we have exactly $(b - a) \div n$ numbers in $[2^W \Mod n,2^W)$ with remainder $r$, thus having a uniform distribution over $[0,n)$ when computing the remainder.
+Due to \hyperref[lemma:1]{Lemma 1}, for every $r \in [0,n)$, we have exactly $(2^W - 2^W \Mod n) \div n$ numbers in $[2^W \Mod n,2^W)$ with remainder $r$, thus having a uniform distribution over $[0,n)$ when computing the remainder.
 This yields \hyperref[alg:openbsd]{Algorithm 3} which stems from the C standard library in OpenBSD and macOS (\texttt{arc4random\_uniform}), and has been adopted by the Go language (\texttt{Int63n, Int31n}) with slight implementation differences~\cite{GoLang} as well as the GNU C++ standard library~\cite{GnuCpp}.
@@
 
-\noindent Note that the remainder operation in line $3$ and $9$ can be computed by a bitwise \textsc{And} operation and the division operation in line $10$ by a Right-\textsc{BitShift} as mentioned in \hyperref[sec:1.2]{Section 1}
+\newpage % In the final version, this paragraph started on the previous page before the algorithm which could be confusing
+\noindent Note that the remainder operation in line $3$ and $9$ can be computed by a bitwise \textsc{And} operation and the division operation in line $10$ by a Right-\textsc{BitShift} as mentioned in \hyperref[sec:1.2]{Section 1}.
 Thus, the only \emph{real} division operation is the remainder operation in line $5$.

### 2024-02-06 Presentation/preliminaries.tex
presentation structure + motivation + preliminaries

\section{Preliminaries}

\subsection{Formal Definition}
\begin{frame}{Formal Definition}
    \pause
    \onslide<+->{Setting:}
    \begin{itemize}[<+->]
        \item \textbf{Input:} upper bound of interval $n \in \mathbb{N}$
        \item \textbf{Output:} uniform random integer in interval $[0,n)$
    \end{itemize}

    \vspace*{0.5cm}

    \begin{alertblock}{}<+->
        \centering
        But what if we want a random integer in $[a,b)$ for $a,b \in \mathbb{N},\,0 < a < b$ instead?
    \end{alertblock}

    \smallskip

    \onslide<+->{We can map this to our setting by subtracting $a$!}
    \begin{itemize}[<+->]
        \item Set $n = b - a$ and draw a uniform random integer $x \in [0,n)$
        \item Return $x + a$
    \end{itemize}
\end{frame}

\subsection{Operations}
\begin{frame}{Operations}
    \pause 
    \begin{definition}[Common Operations]<+->
        \begin{itemize}[<+->]
            \item \makebox[4cm][l]{Integer-Division:} \makebox[1.5cm][l]{$x \div y$} $\coloneqq \lfloor\nicefrac{x}{y}\rfloor$
            \item \makebox[4cm][l]{Remainder-Operation:} \makebox[1.5cm][l]{$x \Mod y$} $\coloneqq x - (x \div y)y$
            \item \makebox[4cm][l]{Bit-\textsc{RightShift}:} \makebox[1.5cm][l]{$x \gg W$} $\coloneqq x \div 2^W$
            \item \makebox[4cm][l]{Bit-\textsc{LeftShift}:} \makebox[1.5cm][l]{$x \ll W$} $\coloneqq x \cdot 2^W$
            \item \makebox[4cm][l]{Bitwise-\textsc{And}:} \makebox[1.5cm][l]{$x \And y$} \onslide<+->{$\rightarrow x \Mod 2^W \coloneqq x \And (2^W - 1)$} 
        \end{itemize}
    \end{definition}

\end{frame}

\subsection{The Naive Approach}
\begin{frame}{The Naive Approach}
    \pause 
    \begin{block}{How do we get random numbers?}<+->
        \begin{itemize}[<+->]
            \item Generated by Pseudo-Random-Number-Generators (PRNGs)
            \item Generated as $W$-bit words, i.e. unsigned integers in $[0,2^W)$ (typically $W \in \{32,64\}$)
        \end{itemize}
    \end{block}

    \begin{exampleblock}{}<+->
        \begin{center}
            $\texttt{rand()} \Mod n$
        \end{center}
    \end{exampleblock}

    \onslide<+->{Does this work?}
    \begin{itemize}[<+->]
        \item \GB{Yes}, the generated number is in $[0,n)$.
    \end{itemize}

    \onslide<+->{Is this efficient?}
    \begin{itemize}[<+->]
        \item \ER{No}, we require one expensive integer division operation.
    \end{itemize}

    \onslide<+->{Is the generated number uniform in $[0,n)$?}
\end{frame}

\iffalse
\begin{frame}{The Naive Approach}
    \begin{alertblock}{Is the generated number uniform in $\boldsymbol{[0,n)}$?}<+->
        \onslide<+->{Consider the case where $W = 4$ and $n = 3$.}
        \onslide<+->{Then, $[0,2^W)$ is}
        \onslide<+->{\begin{align*}
            0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15
        \end{align*}}
        \onslide<+->{Apply $\left(x \Mod 3\right)$ for every $x$ and get}
        \onslide<+->{\begin{align*}
            0,1,2,0,1,2,0,1,2,0,1,2,0,1,2,0
        \end{align*}}
        \onslide<+->{\GB{$0$} appears \ER{$6$} times, \GB{$1$} appears \ER{$5$} times, and \GB{$2$} appears \ER{$5$} times!}
    \end{alertblock}

\end{frame}
\fi

\begin{frame}{The Naive Approach}
    \pause 
    \onslide<+->{In general, applying $\left(x \Mod n\right)$ to $[0,2^W)$ yields}
    \onslide<+->{\begin{align*}
        \overbrace{\underbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\underbrace{0,1,\ldots,\left(2^W \Mod n\right) - 1}_{\text{$2^W \Mod n$ values}}}}^{\text{$2^W$ values}}
    \end{align*}}

    \onslide<+->{We have a \ER{leftover} interval that introduces bias.}

    \vspace*{0.5cm}

    \onslide<+->{\GB{Every} approach that maps every integer in $[0,2^W)$ to a single number in $[0,n)$}
    \onslide<+->{does \ER{not} generate uniform random integers in one step.}

\end{frame}

### 2024-02-06 Presentation/unbiased.tex
presentation structure + motivation + preliminaries

\section{Unbiased Algorithms}

\subsection{The OpenBSD Algorithm}
\begin{frame}{The OpenBSD Algorithm}
    \pause

\end{frame}

\subsection{The Java Algorithm}
\begin{frame}{The Java Algorithm}
    \pause

\end{frame}

\subsection{The Fast-Dice-Roller Algorithm}
\begin{frame}{The Fast-Dice-Roller Algorithm}
    \pause

\end{frame}

\subsection{The Bitmask Algorithm}
\begin{frame}{The Bitmask Algorithm}
    \pause

\end{frame}

### 2024-02-06 Presentation/preliminaries.tex
color in math mode + OpenBSD

@@
 
+    \vspace*{0.5cm}
+
+    \begin{definition}[Power Remainder]<+->
+        For $W,n \in \mathbb{N}$, we write $\RNW$ for $2^W \Mod n$.
+    \end{definition}
+
 \end{frame}
@@
 
-    \begin{exampleblock}{}<+->
-        \begin{center}
+    \onslide<+->{\begin{center}
             $\texttt{rand()} \Mod n$
-        \end{center}
-    \end{exampleblock}
+    \end{center}}
 
@@
 
-\iffalse
-\begin{frame}{The Naive Approach}
-    \begin{alertblock}{Is the generated number uniform in $\boldsymbol{[0,n)}$?}<+->
-        \onslide<+->{Consider the case where $W = 4$ and $n = 3$.}
-        \onslide<+->{Then, $[0,2^W)$ is}
-        \onslide<+->{\begin{align*}
-            0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15
-        \end{align*}}
-        \onslide<+->{Apply $\left(x \Mod 3\right)$ for every $x$ and get}
-        \onslide<+->{\begin{align*}
-            0,1,2,0,1,2,0,1,2,0,1,2,0,1,2,0
-        \end{align*}}
-        \onslide<+->{\GB{$0$} appears \ER{$6$} times, \GB{$1$} appears \ER{$5$} times, and \GB{$2$} appears \ER{$5$} times!}
-    \end{alertblock}
-
-\end{frame}
-\fi
-
 \begin{frame}{The Naive Approach}
     \pause 
-    \onslide<+->{In general, applying $\left(x \Mod n\right)$ to $[0,2^W)$ yields}
-    \onslide<+->{\begin{align*}
-        \overbrace{\underbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\underbrace{0,1,\ldots,\left(2^W \Mod n\right) - 1}_{\text{$2^W \Mod n$ values}}}}^{\text{$2^W$ values}}
+    \onslide<+->{In general, applying $x \Mod n$ to $[0,2^W)$ yields}
+    \onslide<+->{\begin{align*}\everymath{} % Allows specific coloring
+        \overbrace{\underbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\underbrace{0,1,\ldots,\RNW - 1}_{\text{$\RNW$ values}}}}^{\text{$2^W$ values}}
     \end{align*}}
@@
     \onslide<+->{\GB{Every} approach that maps every integer in $[0,2^W)$ to a single number in $[0,n)$}
-    \onslide<+->{does \ER{not} generate uniform random integers in one step.}
+    \onslide<+->{does \ER{not} generate uniform random integers in one step}
+    \onslide<+->{whenever $n$ does not divide $2^W$.}
 
+    \vspace*{0.5cm}
+
+    \onslide<+->{\GB{Idea:} Use \ER{rejection sampling} to achieve uniformity!}
 \end{frame}

### 2024-02-08 Presentation/main.tex
.
presentation structure + motivation + preliminaries
color in math mode + OpenBSD
java algorithm

@@
 \RequirePackage{marvosym}
+
+\usepackage{mathtools}
+\usepackage{nicefrac}
+\newcommand{\Mod}[1]{\ \mathrm{mod}\ #1}
+
+\newcommand{\GB}[1]{\textcolor{goetheblau}{#1}}
+\newcommand{\ER}[1]{\textcolor{emorot}{#1}}
+\newcommand{\DG}[1]{\textcolor{dunkelgruen}{#1}}
+\newcommand{\BB}[1]{\textcolor{black}{#1}}
+
+\newcommand{\RNW}{\ensuremath{\mathcal{R}_n^W}}
+
+\usepackage{everysel}
+\everymath{\color{goetheblau}}
 
@@
 
+\tikzstyle{overbrace text style}=[font=\small, above, pos=.5, yshift=3mm]
+\tikzstyle{overbrace style}=[decorate,decoration={brace,raise=2mm,amplitude=3pt}]
+\tikzstyle{underbrace style}=[decorate,decoration={brace,raise=2mm,amplitude=3pt,mirror},color=gray]
+\tikzstyle{underbrace text style}=[font=\small, below, pos=.5, yshift=-3mm]
+
+\usepackage{cancel}
+
 \subject{Seminar Algorithms for Big Data}
@@
 \author{Lukas Geis}
-\supervisor{Supervised by Dr Manuel Penschuck}
+\supervisor{Supervised by Dr. Manuel Penschuck}
 \date{29th February 2024}
-\institute{Algorithm Engineering (Prof. Dr Ulrich Meyer)}
+\institute{Algorithm Engineering (Prof. Dr. Ulrich Meyer)}
 
@@
 
+	\include{motivation}
+
 	\begin{frame}{Table of Contents}
@@
 	\end{frame}
+
+	\include{preliminaries}
+
+	\include{unbiased}
+
+	\include{lemire}
+
+	\include{conclusion}
 \end{document}

### 2024-02-08 Presentation/unbiased.tex
color in math mode + OpenBSD
java algorithm

@@
     \pause 
+    \begin{itemize}[<+->]
+        \item We shift the \ER{rejection interval} to the left:
+        \onslide<+->{\begin{align*}\everymath{} % Allows specific coloring
+            \overbrace{\ER{\underbrace{0,1,\ldots,\RNW - 1}_{\text{$\RNW$ values}}},\underbrace{\overbrace{\RNW,\ldots,n - 1,0,\ldots,\RNW - 1}^{\text{$n$ values}},\ldots,\overbrace{\RNW,\ldots,n - 1,0,\ldots,\RNW - 1}^{\text{$n$ values}}}_{\text{$\left(2^W \div n\right) \cdot n$ values}}}^{\text{$2^W$ values}}
+        \end{align*}}
+        \item Generate a uniform random number $x \in [0,2^W)$ until $x \geq \RNW$
+        \item Return $x \Mod n$
+    \end{itemize}
+
+    \begin{block}{Efficiency}<+->
+        \onslide<+->{We require $2$ integer division operations:}
+        \onslide<+->{one for computing $\RNW$}
+        \onslide<+->{and one for computing $x \Mod n$.}
+    \end{block}
 
 \end{frame}
+
+%%% Array Drawing with Tikz instead of Equation
+\iffalse
+\onslide<+->{\begin{center}
+    \begin{tikzpicture}[scale = 0.7]
+        \draw[draw = goetheblau] (0.0,0.0) rectangle ++(5.0,1.0);
+        \draw[draw = goetheblau,overbrace style] (0.0,1.0) -- (5.0,1.0) node [overbrace text style] {\GB{$n$ values}};
+        \draw[draw = goetheblau] (5.0,0.0) rectangle ++(5.0,1.0);
+        \draw[draw = goetheblau,overbrace style] (5.0,1.0) -- (10.0,1.0) node [overbrace text style] {\GB{$n$ values}};
+        \node at (10.5,0.5) {$\cdots$};
+        \draw[draw = goetheblau] (11.0,0.0) rectangle ++(5.0,1.0);
+        \draw[draw = goetheblau,overbrace style] (11.0,1.0) -- (16.0,1.0) node [overbrace text style] {\GB{$n$ values}};
+        \draw[draw = emorot] (16.0,0.0) rectangle ++(3.0,1.0);
+        \draw[draw = emorot,overbrace style] (16.0,1.0) -- (19.0,1.0) node [overbrace text style] {$\text{\ER{\RNW values}}$};
+    \end{tikzpicture}
+\end{center}}
+\fi
 
@@
     \pause 
+    \begin{itemize}
+        \item<1-> Consider $x - \left(x \Mod n\right)$ for $x \in [0,2^W)$: \only<3>{\begin{align*}\everymath{} % Allows specific coloring
+            \overbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
+        \end{align*}} \only<4>{\begin{align*}\everymath{} % Allows specific coloring
+            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
+        \end{align*}} \only<5>{\begin{align*}\everymath{} % Allows specific coloring
+            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
+        \end{align*}} \only<6>{\begin{align*}\everymath{} % Allows specific coloring
+            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\substack{\text{mapped to} \\ 2^W - n - \RNW}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
+        \end{align*}} \only<7->{\begin{align*}\everymath{} % Allows specific coloring
+            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\substack{\text{mapped to} \\ 2^W - n - \RNW}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\underbrace{\overbrace{\BB{0},1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}_{\substack{\text{mapped to} \\ 2^W - \RNW}}}
+        \end{align*}}
+        \item<8-> Only numbers in the \ER{leftover} interval get mapped to $\ER{2^W - \RNW} > 2^W - n$
+        \item<9-> Algorithm: \begin{itemize}
+            \item[(1)]<10-> Draw $x \in [0,2^W)$ and compute $r = x \Mod n$
+            \item[(2)]<11-> Return $r$ if $x - r > 2^W - n$ else goto \textbf{\GB{(1)}}
+        \end{itemize}
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{The Java Algorithm}
+    Algorithm: \begin{itemize}
+        \item[(1)] Draw $x \in [0,2^W)$ and compute $r = x \Mod n$
+        \item[(2)] Return $r$ if $x - r > 2^W - n$ else goto \textbf{\GB{(1)}}
+    \end{itemize}
+    \pause 
+    \begin{block}{Efficiency}<+->
+        \begin{itemize}[<+->]
+            \item At least one integer division operation
+            \item Number of integer divisions operations equal to number of \GB{rounds}
+            \item Return a number in \GB{round} if $x < 2^W - \RNW$
+            \item Happens with probability $\frac{2^W - \RNW}{2^W} > \frac{1}{2}$
+            \item Expected number of integer division operations is $\frac{2^W}{2^W - \RNW} < 2$
+        \end{itemize}
+    \end{block}
 

### 2024-02-11 Presentation/unbiased.tex
remove FDR + summary table (need to fix vertical line)

@@
 
-\subsection{The OpenBSD Algorithm}
+\subsection{The OpenBSD Algorithm}\label{sec:2.1}
 \begin{frame}{The OpenBSD Algorithm}
@@
 
-%%% Array Drawing with Tikz instead of Equation
-\iffalse
-\onslide<+->{\begin{center}
-    \begin{tikzpicture}[scale = 0.7]
-        \draw[draw = goetheblau] (0.0,0.0) rectangle ++(5.0,1.0);
-        \draw[draw = goetheblau,overbrace style] (0.0,1.0) -- (5.0,1.0) node [overbrace text style] {\GB{$n$ values}};
-        \draw[draw = goetheblau] (5.0,0.0) rectangle ++(5.0,1.0);
-        \draw[draw = goetheblau,overbrace style] (5.0,1.0) -- (10.0,1.0) node [overbrace text style] {\GB{$n$ values}};
-        \node at (10.5,0.5) {$\cdots$};
-        \draw[draw = goetheblau] (11.0,0.0) rectangle ++(5.0,1.0);
-        \draw[draw = goetheblau,overbrace style] (11.0,1.0) -- (16.0,1.0) node [overbrace text style] {\GB{$n$ values}};
-        \draw[draw = emorot] (16.0,0.0) rectangle ++(3.0,1.0);
-        \draw[draw = emorot,overbrace style] (16.0,1.0) -- (19.0,1.0) node [overbrace text style] {$\text{\ER{\RNW values}}$};
-    \end{tikzpicture}
-\end{center}}
-\fi
-
-\subsection{The Java Algorithm}
+\subsection{The Java Algorithm}\label{sec:2.2}
 \begin{frame}{The Java Algorithm}
@@
         \end{align*}}
-        \item<8-> Only numbers in the \ER{leftover} interval get mapped to $\ER{2^W - \RNW} > 2^W - n$
-        \item<9-> Algorithm: \begin{itemize}
-            \item[(1)]<10-> Draw $x \in [0,2^W)$ and compute $r = x \Mod n$
-            \item[(2)]<11-> Return $r$ if $x - r > 2^W - n$ else goto \textbf{\GB{(1)}}
+        \item<8-> We map every number to the next-smallest multiple of $n$
+        \item<9-> Only numbers in the \ER{leftover} interval get mapped to $\ER{2^W - \RNW} > 2^W - n$
+        \item<10-> Algorithm: \begin{itemize}
+            \item[(1)]<11-> Draw $x \in [0,2^W)$ and compute $r = x \Mod n$
+            \item[(2)]<12-> Return $r$ if $x - r > 2^W - n$ else goto \textbf{\GB{(1)}}
         \end{itemize}
@@
 
-\subsection{The Fast-Dice-Roller Algorithm}
-\begin{frame}{The Fast-Dice-Roller Algorithm}
+\subsection{The Bitmask Algorithm}\label{sec:2.3}
+\begin{frame}{The Bitmask Algorithm}
     \pause
@@
 
-\subsection{The Bitmask Algorithm}
-\begin{frame}{The Bitmask Algorithm}
+%%% Array Drawing with Tikz instead of Equation + FastDiceRoller
+\iffalse
+\onslide<+->{\begin{center}
+    \begin{tikzpicture}[scale = 0.7]
+        \draw[draw = goetheblau] (0.0,0.0) rectangle ++(5.0,1.0);
+        \draw[draw = goetheblau,overbrace style] (0.0,1.0) -- (5.0,1.0) node [overbrace text style] {\GB{$n$ values}};
+        \draw[draw = goetheblau] (5.0,0.0) rectangle ++(5.0,1.0);
+        \draw[draw = goetheblau,overbrace style] (5.0,1.0) -- (10.0,1.0) node [overbrace text style] {\GB{$n$ values}};
+        \node at (10.5,0.5) {$\cdots$};
+        \draw[draw = goetheblau] (11.0,0.0) rectangle ++(5.0,1.0);
+        \draw[draw = goetheblau,overbrace style] (11.0,1.0) -- (16.0,1.0) node [overbrace text style] {\GB{$n$ values}};
+        \draw[draw = emorot] (16.0,0.0) rectangle ++(3.0,1.0);
+        \draw[draw = emorot,overbrace style] (16.0,1.0) -- (19.0,1.0) node [overbrace text style] {$\text{\ER{\RNW values}}$};
+    \end{tikzpicture}
+\end{center}}
+
+\subsection{The Fast-Dice-Roller Algorithm}
+\begin{frame}{The Fast-Dice-Roller Algorithm}
     \pause 
+    \begin{itemize}[<+->]
+        \item Build up $x$ bit-by-bit using uniform random bits \GB{\texttt{flip()}}
+        \item Keep track of upper bound $\mathcal{B}$ for number \onslide<+->{  $\longrightarrow \quad x \in [0,\mathcal{B})$}
+        \item Repeat until $\mathcal{B} \geq n$ \begin{itemize}[<+->]
+            \item if $x < n$, return $x$
+            \item else decrease $x$ and $\mathcal{B}$ by $n$ (\ER{rejection})
+        \end{itemize}
+    \end{itemize}
 
 \end{frame}
+\fi

### 2024-02-12 Presentation/unbiased.tex
a bit of Bitmask - will finish later

@@
         \end{align*}}
+        \item Algorithm: \begin{itemize}[<+->]
             \item Generate a uniform random number $x \in [0,2^W)$ until $x \geq \RNW$
@@
         \end{itemize} 
+    \end{itemize}
 
+\end{frame}
+
+\begin{frame}{The OpenBSD Algorithm}
+    Algorithm: \begin{itemize}
+        \item Generate a uniform random number $x \in [0,2^W)$ until $x \geq \RNW$
+        \item Return $x \Mod n$
+    \end{itemize} 
+    \pause 
+    \smallskip
     \begin{block}{Efficiency}<+->
-        \onslide<+->{We require $2$ integer division operations:}
-        \onslide<+->{one for computing $\RNW$}
-        \onslide<+->{and one for computing $x \Mod n$.}
+        \onslide<+->{We require $2$ integer division operations:} \begin{itemize}[<+->]
+            \item one for computing $\RNW$
+            \item and one for computing $x \Mod n$.
+        \end{itemize}
     \end{block}
-
+    \vspace*{0.5cm} 
+    \onslide<+->{Can we do better?}
 \end{frame}
@@
     \pause 
+    \smallskip
     \begin{block}{Efficiency}<+->
@@
     \end{block}
-
 \end{frame}
@@
 \subsection{The Bitmask Algorithm}\label{sec:2.3}
+\begin{frame}{The Bitmask Algorithm}
+    \pause 
+    \begin{itemize}[<+->]
+        \item Consider the \GB{binary} representation of $n$: \onslide<+->{\begin{align*}\everymath{}
+            \underbrace{
+                \overbrace{0}^{
+                    \hidewidth 2^{W - 1} \hidewidth 
+                },\ldots,0
+            }_{\text{only $0$'s}},\ER{\overbrace{1}^{
+                 2^{\lfloor\log_2n\rfloor}
+            }},\underbrace{
+                \nicefrac{0}{1},\ldots,\overbrace{
+                    \nicefrac{0}{1}
+                }^{ 2^1 },\overbrace{
+                    \nicefrac{0}{1}
+                }^{ 2^0 }
+            }_{\text{series of $0$'s and $1$'s}}
+        \end{align*}} 
+        \item Every number $x \leq n$ only needs the last $\lfloor\log_2n\rfloor + 1$ bits
+        \item Get these bits with a bitwise-\textsc{And} with $2^{\lfloor\log_2n\rfloor + 1} - 1$: \onslide<+->{\begin{align*}\everymath{}
+            \underbrace{
+                \overbrace{0}^{
+                    \hidewidth 2^{W - 1} \hidewidth 
+                },\ldots,0
+            }_{\text{only $0$'s}},\underbrace{\ER{\overbrace{1}^{
+                 2^{\lfloor\log_2n\rfloor}
+            }},1,\ldots,\overbrace{
+                    1
+                }^{ 2^1 },\overbrace{
+                    1
+                }^{ 2^0 }
+            }_{\text{only $1$'s}}
+        \end{align*}}
+        \item Every resulting number $x \And \left(2^{\lfloor\log_2n\rfloor + 1} - 1\right)$ is at most \onslide<+->{$2^{\lfloor\log_2n\rfloor + 1} - 1 < 2n$}
+    \end{itemize}
+\end{frame}
+
 \begin{frame}{The Bitmask Algorithm}

### 2024-02-12 Presentation/unbiased.tex
Bitmask

@@
         \item Consider the \GB{binary} representation of $n$: \onslide<+->{\begin{align*}\everymath{}
-            \underbrace{
+            n \quad \stackrel{\text{binary}}{\longrightarrow} \qquad \underbrace{
                 \overbrace{0}^{
@@
         \item Every number $x \leq n$ only needs the last $\lfloor\log_2n\rfloor + 1$ bits
-        \item Get these bits with a bitwise-\textsc{And} with $2^{\lfloor\log_2n\rfloor + 1} - 1$: \onslide<+->{\begin{align*}\everymath{}
-            \underbrace{
+        \item Get these bits with a bitwise-\textsc{And} with \begin{align*}\everymath{}
+            2^{\lfloor\log_2n\rfloor + 1} - 1 \quad \stackrel{\text{binary}}{\longrightarrow} \qquad \underbrace{
                 \overbrace{0}^{
@@
             }_{\text{only $1$'s}}
-        \end{align*}}
-        \item Every resulting number $x \And \left(2^{\lfloor\log_2n\rfloor + 1} - 1\right)$ is at most \onslide<+->{$2^{\lfloor\log_2n\rfloor + 1} - 1 < 2n$}
+        \end{align*}
     \end{itemize}
@@
     \pause 
+    \begin{itemize}[<+->]
+        \item How can we compute $2^{\lfloor\log_2n\rfloor + 1}$? 
+        \item \GB{Count} the number $\ell$ of \GB{leading $0$'s}! \onslide<+->{\begin{align*}\everymath{}
+            n \quad \stackrel{\text{binary}}{\longrightarrow} \qquad \underbrace{
+                \overbrace{0}^{
+                    \hidewidth 2^{W - 1} \hidewidth 
+                },\ldots,0
+            }_{\text{$\ell$ $0$'s}},\ER{\overbrace{1}^{
+                 2^{\lfloor\log_2n\rfloor}
+            }},\underbrace{
+                \nicefrac{0}{1},\ldots,\overbrace{
+                    \nicefrac{0}{1}
+                }^{ 2^1 },\overbrace{
+                    \nicefrac{0}{1}
+                }^{ 2^0 }
+            }_{\text{series of $0$'s and $1$'s}}
+        \end{align*}} 
+        \item $\ER{\lfloor\log_2n\rfloor} = W - \ell - 1$ \onslide<+->{$\qquad \longrightarrow \qquad 2^{\lfloor\log_2n\rfloor + 1} = 1 \ll \left(W - \ell\right)$}
+        \item Algorithm: \begin{itemize}[<+->]
+            \item[(1)] Compute $\ell$ and then $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
+            \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
+            \item[(3)] Return $b$ if $b < n$ else goto \textbf{\GB{(2)}}
+        \end{itemize}
+    \end{itemize}
+\end{frame}
 
+\begin{frame}{The Bitmask Algorithm}
+    Algorithm: \begin{itemize}
+        \item[(1)] Compute $\ell$ and then $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
+        \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
+        \item[(3)] Return $b$ if $b < n$ else goto \textbf{\GB{(2)}}
+    \end{itemize}
+    \pause 
+    \smallskip
+    \begin{block}{Efficiency}<+->
+        \begin{itemize}[<+->]
+            \item $b$ is at most $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1 < 2n$ \onslide<+->{$\quad \longrightarrow \quad$ success probability at least $\approx \frac{1}{2}$}
+            \item At most $2$ rounds in expectation
+            \item \GB{No} integer division at all
+            \item Computation of \GB{leading $0$'s} needs \GB{\texttt{clz}} instruction 
+            \item Roughly as expensive as a \GB{\texttt{div}} instruction 
+        \end{itemize}
+    \end{block}
 \end{frame}

### 2024-02-12 Presentation/unbiased.tex
shorter sentences

@@
         \item Algorithm: \begin{itemize}[<+->]
-            \item[(1)] Compute $\ell$ and then $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
+            \item[(1)] Compute $\ell$ and $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
             \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
@@
     Algorithm: \begin{itemize}
-        \item[(1)] Compute $\ell$ and then $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
+        \item[(1)] Compute $\ell$ and $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
         \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
@@
         \begin{itemize}[<+->]
-            \item $b$ is at most $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1 < 2n$ \onslide<+->{$\quad \longrightarrow \quad$ success probability at least $\approx \frac{1}{2}$}
+            \item $b$ at most $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1 < 2n$ \onslide<+->{$\quad \longrightarrow \quad$ success probability at least $\approx \frac{1}{2}$}
             \item At most $2$ rounds in expectation
             \item \GB{No} integer division at all
-            \item Computation of \GB{leading $0$'s} needs \GB{\texttt{clz}} instruction 
+            \item Computation of \GB{leading $0$'s} requires \GB{\texttt{clz}} instruction/algorithm 
             \item Roughly as expensive as a \GB{\texttt{div}} instruction 

### 2024-02-12 Presentation/unbiased.tex
Multiply-And-Shift

@@
         \end{align*}}
-        \item<8-> We map every number to the next-smallest multiple of $n$
-        \item<9-> Only numbers in the \ER{leftover} interval get mapped to $\ER{2^W - \RNW} > 2^W - n$
+        \item<8-> Map every number to the next-smallest multiple of $n$
+        \item<9-> Only numbers in \ER{leftover} interval mapped to $\ER{2^W - \RNW} > 2^W - n$
         \item<10-> Algorithm: \begin{itemize}
@@
             \item Number of integer divisions operations equal to number of \GB{rounds}
-            \item Return a number in \GB{round} if $x < 2^W - \RNW$
+            \item Return number in \GB{round} if $x < 2^W - \RNW$
             \item Happens with probability $\frac{2^W - \RNW}{2^W} > \frac{1}{2}$

### 2024-02-13 Presentation/unbiased.tex
some fixes

@@
     \begin{itemize}[<+->]
-        \item We shift the \ER{rejection interval} to the left:
+        \item Shift the \ER{rejection interval} to the left:
         \onslide<+->{\begin{align*}\everymath{} % Allows specific coloring
@@
     \pause 
-    \smallskip
+    \vspace*{0.5cm}
     \begin{block}{Efficiency}<+->
@@
     \end{block}
-    \vspace*{0.5cm} 
-    \onslide<+->{Can we do better?}
 \end{frame}
@@
         \item Consider the \GB{binary} representation of $n$: \onslide<+->{\begin{align*}\everymath{}
-            n \quad \stackrel{\text{binary}}{\longrightarrow} \qquad \underbrace{
+            n \quad \xrightarrow{\text{binary}} \qquad \underbrace{
                 \overbrace{0}^{
@@
         \item Get these bits with a bitwise-\textsc{And} with \begin{align*}\everymath{}
-            2^{\lfloor\log_2n\rfloor + 1} - 1 \quad \stackrel{\text{binary}}{\longrightarrow} \qquad \underbrace{
+            2^{\lfloor\log_2n\rfloor + 1} - 1 \quad \xrightarrow{\text{binary}} \qquad \underbrace{
                 \overbrace{0}^{
@@
         \item \GB{Count} the number $\ell$ of \GB{leading $0$'s}! \onslide<+->{\begin{align*}\everymath{}
-            n \quad \stackrel{\text{binary}}{\longrightarrow} \qquad \underbrace{
+            n \quad \xrightarrow{\text{binary}} \qquad \underbrace{
                 \overbrace{0}^{
@@
             \item $b$ at most $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1 < 2n$ \onslide<+->{$\quad \longrightarrow \quad$ success probability at least $\approx \frac{1}{2}$}
-            \item At most $2$ rounds in expectation
+            \item At most $\approx 2$ rounds in expectation
             \item \GB{No} integer division at all

### 2024-02-15 Presentation/unbiased.tex
finished presentation v1

@@
 
-\begin{frame}{The OpenBSD Algorithm}
+\begin{frame}{The OpenBSD Algorithm - Efficiency}
     Algorithm: \begin{itemize}
@@
             \item[(1)]<11-> Draw $x \in [0,2^W)$ and compute $r = x \Mod n$
-            \item[(2)]<12-> Return $r$ if $x - r > 2^W - n$ else goto \textbf{\GB{(1)}}
+            \item[(2)]<12-> Return $r$ if $x - r \leq 2^W - n$ else goto \textbf{\GB{(1)}}
         \end{itemize}
@@
 
-\begin{frame}{The Java Algorithm}
+\begin{frame}{The Java Algorithm - Efficiency}
     Algorithm: \begin{itemize}
         \item[(1)] Draw $x \in [0,2^W)$ and compute $r = x \Mod n$
-        \item[(2)] Return $r$ if $x - r > 2^W - n$ else goto \textbf{\GB{(1)}}
+        \item[(2)] Return $r$ if $x - r \leq 2^W - n$ else goto \textbf{\GB{(1)}}
     \end{itemize}
@@
 \subsection{The Bitmask Algorithm}\label{sec:2.3}
-\begin{frame}{The Bitmask Algorithm}
+\begin{frame}{The Bitmask Algorithm - Representation}
     \pause 
@@
 
-\begin{frame}{The Bitmask Algorithm}
+\begin{frame}{The Bitmask Algorithm - Mask}
     \pause 
@@
         \item How can we compute $2^{\lfloor\log_2n\rfloor + 1}$? 
-        \item \GB{Count} the number $\ell$ of \GB{leading $0$'s}! \onslide<+->{\begin{align*}\everymath{}
+        \item \GB{Count} the number $\ell$ of \GB{leading $0$'s} in $n$! \onslide<+->{\begin{align*}\everymath{}
             n \quad \xrightarrow{\text{binary}} \qquad \underbrace{
@@
 
-\begin{frame}{The Bitmask Algorithm}
+\begin{frame}{The Bitmask Algorithm - Efficiency}
     Algorithm: \begin{itemize}
@@
 \end{frame}
-
-%%% Array Drawing with Tikz instead of Equation + FastDiceRoller
-\iffalse
-\onslide<+->{\begin{center}
-    \begin{tikzpicture}[scale = 0.7]
-        \draw[draw = goetheblau] (0.0,0.0) rectangle ++(5.0,1.0);
-        \draw[draw = goetheblau,overbrace style] (0.0,1.0) -- (5.0,1.0) node [overbrace text style] {\GB{$n$ values}};
-        \draw[draw = goetheblau] (5.0,0.0) rectangle ++(5.0,1.0);
-        \draw[draw = goetheblau,overbrace style] (5.0,1.0) -- (10.0,1.0) node [overbrace text style] {\GB{$n$ values}};
-        \node at (10.5,0.5) {$\cdots$};
-        \draw[draw = goetheblau] (11.0,0.0) rectangle ++(5.0,1.0);
-        \draw[draw = goetheblau,overbrace style] (11.0,1.0) -- (16.0,1.0) node [overbrace text style] {\GB{$n$ values}};
-        \draw[draw = emorot] (16.0,0.0) rectangle ++(3.0,1.0);
-        \draw[draw = emorot,overbrace style] (16.0,1.0) -- (19.0,1.0) node [overbrace text style] {$\text{\ER{\RNW values}}$};
-    \end{tikzpicture}
-\end{center}}
-
-\subsection{The Fast-Dice-Roller Algorithm}
-\begin{frame}{The Fast-Dice-Roller Algorithm}
-    \pause 
-    \begin{itemize}[<+->]
-        \item Build up $x$ bit-by-bit using uniform random bits \GB{\texttt{flip()}}
-        \item Keep track of upper bound $\mathcal{B}$ for number \onslide<+->{  $\longrightarrow \quad x \in [0,\mathcal{B})$}
-        \item Repeat until $\mathcal{B} \geq n$ \begin{itemize}[<+->]
-            \item if $x < n$, return $x$
-            \item else decrease $x$ and $\mathcal{B}$ by $n$ (\ER{rejection})
-        \end{itemize}
-    \end{itemize}
-
-\end{frame}
-\fi

### 2024-02-25 Presentation/main.tex
remove FDR + summary table (need to fix vertical line)
table fix
finished presentation v1
new motivation + better [a,b) visual + bit representation + further fixes

@@
 
-\newcommand{\RNW}{\ensuremath{\mathcal{R}_n^W}}
+\newcommand{\RNW}{\ensuremath{\mathcal{R}}}
+
+\usepackage{array}
+\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}} % Centered Multilines
+\usepackage{booktabs}
+\usepackage{pifont} % checkmark / cross - thanks to `https://tex.stackexchange.com/questions/42619/xmark-that-complements-the-ams-checkmark`
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
 
@@
 \everymath{\color{goetheblau}}
+
+\usepackage[mode=buildnew]{standalone}
 
@@
 	every path/.style = {line width=.57mm, line cap=round},
+	time/.style={very thick,|-latex, gray},
+	int/.style={ultra thick,{|[scale=.5]}-{|[scale=.5]}},
 }
@@
 \tikzstyle{underbrace text style}=[font=\small, below, pos=.5, yshift=-3mm]
+
+\usepackage{xparse}
+
+% Thanks to `https://tex.stackexchange.com/questions/8720/overbrace-underbrace-but-with-an-arrow-instead`
+\renewcommand{\arraystretch}{0.7}
+\NewDocumentCommand{\overarrow}{O{0} O{\big\downarrow} m}{%
+  \overset{\makebox[0pt]{\begin{tabular}{@{}c@{}}$#3$\\[0pt]\ensuremath{#2}\\[0pt]\hfil\end{tabular}}}{#1}
+}
 

### 2024-02-25 Presentation/preliminaries.tex
remove FDR + summary table (need to fix vertical line)
some fixes
finished presentation v1
new motivation + better [a,b) visual + bit representation + further fixes

@@
     \pause
-    \onslide<+->{Setting:}
-    \begin{itemize}[<+->]
-        \item \textbf{Input:} upper bound of interval $n \in \mathbb{N}$
-        \item \textbf{Output:} uniform random integer in interval $[0,n)$
+    \onslide<+->{\textbf{Input:}} \begin{itemize}[<+->]
+        \item source of uniform random integers in $[0,2^W)$: \texttt{rand()}
+        \item upper bound of interval $n \in \mathbb{N}$
     \end{itemize}
+    \onslide<+->{\textbf{Output:} \begin{itemize}
+        \item uniform random integer in interval $[0,n)$
+    \end{itemize}}
 
@@
 
-    \begin{alertblock}{}<+->
-        \centering
-        But what if we want a random integer in $[a,b)$ for $a,b \in \mathbb{N},\,0 < a < b$ instead?
-    \end{alertblock}
+    \begin{center}
+        \begin{tikzpicture}[scale = 1.5]
+            \draw<6->[time] (0,0) -- (9,0);
+            \node<6->[time] at (0.015,-0.25) {$0$};
 
-    \smallskip
+            \draw<7->[int, emorot] (5,0) -- (8,0);
+            \node<7-> at (5.015,-0.25) {$\ER{a}$};
+            \node<7-> at (8.0,-0.25) {$\ER{b}$};
+            \node<7-> at (6.5,0.35) {$\ER{[a,b)}$};
 
-    \onslide<+->{We can map this to our setting by subtracting $a$!}
-    \begin{itemize}[<+->]
-        \item Set $n = b - a$ and draw a uniform random integer $x \in [0,n)$
-        \item Return $x + a$
-    \end{itemize}
+            \draw<8->[int, goetheblau] (0,0) -- (3,0);
+            \node<8-> at (3.0,-0.25) {$b - a$};
+            \node<8-> at (1.5,0.35) {$[0,b - a)$};
+
+            \draw<9->[time,black] (3.3,0.5) -- (4.7,0.5);
+            \node<9-> at (4.0,0.72) {$\BB{+\,a}$};
+
+            \node[opacity=0] at (4.5,1.0) {};
+        \end{tikzpicture}
+    \end{center}
 \end{frame}
@@
 
-\subsection{The Naive Approach}
+\subsection{The Naive Approach}\label{sec:1.3}
 \begin{frame}{The Naive Approach}
     \pause 
-    \begin{block}{How do we get random numbers?}<+->
-        \begin{itemize}[<+->]
-            \item Generated by Pseudo-Random-Number-Generators (PRNGs)
-            \item Generated as $W$-bit words, i.e. unsigned integers in $[0,2^W)$ (typically $W \in \{32,64\}$)
-        \end{itemize}
-    \end{block}
-
     \onslide<+->{\begin{center}
-            $\texttt{rand()} \Mod n$
+            \large{$\texttt{rand()} \Mod n$}
     \end{center}}
@@
 
-\begin{frame}{The Naive Approach}
+\begin{frame}{The Naive Approach - Bias}
     \pause 
@@
 
-    \onslide<+->{\GB{Every} approach that maps every integer in $[0,2^W)$ to a single number in $[0,n)$}
-    \onslide<+->{does \ER{not} generate uniform random integers in one step}
+    \begin{block}{Deterministic Mappings}<+->
+        \onslide<+->{\GB{Every} deterministic mapping $f\colon[0,2^W) \rightarrow [0,n)$}
+        \onslide<+->{does \ER{not} generate \GB{uniform} random integers in one step}
         \onslide<+->{whenever $n$ does not divide $2^W$.}
-
-    \vspace*{0.5cm}
-
-    \onslide<+->{\GB{Idea:} Use \ER{rejection sampling} to achieve uniformity!}
+    \end{block}
 \end{frame}

### 2024-02-27 Presentation/unbiased.tex
new motivation + better [a,b) visual + bit representation + further fixes
optional bitmask

@@
 \end{frame}
-
-\subsection{The Bitmask Algorithm}\label{sec:2.3}
-\begin{frame}{The Bitmask Algorithm - Representation}
-    \pause 
-    \begin{itemize}[<+->]
-        \item Consider the \GB{binary} representation of $n$: \onslide<+->{\begin{align*}\everymath{}
-            n \quad \xrightarrow{\text{binary}} \qquad \underbrace{
-                \overbrace{0}^{
-                    \hidewidth 2^{W - 1} \hidewidth 
-                },\ldots,0
-            }_{\text{only $0$'s}},\ER{\overbrace{1}^{
-                 2^{\lfloor\log_2n\rfloor}
-            }},\underbrace{
-                \nicefrac{0}{1},\ldots,\overbrace{
-                    \nicefrac{0}{1}
-                }^{ 2^1 },\overbrace{
-                    \nicefrac{0}{1}
-                }^{ 2^0 }
-            }_{\text{series of $0$'s and $1$'s}}
-        \end{align*}} 
-        \item Every number $x \leq n$ only needs the last $\lfloor\log_2n\rfloor + 1$ bits
-        \item Get these bits with a bitwise-\textsc{And} with \begin{align*}\everymath{}
-            2^{\lfloor\log_2n\rfloor + 1} - 1 \quad \xrightarrow{\text{binary}} \qquad \underbrace{
-                \overbrace{0}^{
-                    \hidewidth 2^{W - 1} \hidewidth 
-                },\ldots,0
-            }_{\text{only $0$'s}},\underbrace{\ER{\overbrace{1}^{
-                 2^{\lfloor\log_2n\rfloor}
-            }},1,\ldots,\overbrace{
-                    1
-                }^{ 2^1 },\overbrace{
-                    1
-                }^{ 2^0 }
-            }_{\text{only $1$'s}}
-        \end{align*}
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{The Bitmask Algorithm - Mask}
-    \pause 
-    \begin{itemize}[<+->]
-        \item How can we compute $2^{\lfloor\log_2n\rfloor + 1}$? 
-        \item \GB{Count} the number $\ell$ of \GB{leading $0$'s} in $n$! \onslide<+->{\begin{align*}\everymath{}
-            n \quad \xrightarrow{\text{binary}} \qquad \underbrace{
-                \overbrace{0}^{
-                    \hidewidth 2^{W - 1} \hidewidth 
-                },\ldots,0
-            }_{\text{$\ell$ $0$'s}},\ER{\overbrace{1}^{
-                 2^{\lfloor\log_2n\rfloor}
-            }},\underbrace{
-                \nicefrac{0}{1},\ldots,\overbrace{
-                    \nicefrac{0}{1}
-                }^{ 2^1 },\overbrace{
-                    \nicefrac{0}{1}
-                }^{ 2^0 }
-            }_{\text{series of $0$'s and $1$'s}}
-        \end{align*}} 
-        \item $\ER{\lfloor\log_2n\rfloor} = W - \ell - 1$ \onslide<+->{$\qquad \longrightarrow \qquad 2^{\lfloor\log_2n\rfloor + 1} = 1 \ll \left(W - \ell\right)$}
-        \item Algorithm: \begin{itemize}[<+->]
-            \item[(1)] Compute $\ell$ and $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
-            \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
-            \item[(3)] Return $b$ if $b < n$ else goto \textbf{\GB{(2)}}
-        \end{itemize}
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{The Bitmask Algorithm - Efficiency}
-    Algorithm: \begin{itemize}
-        \item[(1)] Compute $\ell$ and $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
-        \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
-        \item[(3)] Return $b$ if $b < n$ else goto \textbf{\GB{(2)}}
-    \end{itemize}
-    \pause 
-    \smallskip
-    \begin{block}{Efficiency}<+->
-        \begin{itemize}[<+->]
-            \item $b$ at most $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1 < 2n$ \onslide<+->{$\quad \longrightarrow \quad$ success probability at least $\approx \frac{1}{2}$}
-            \item At most $\approx 2$ rounds in expectation
-            \item \GB{No} integer division at all
-            \item Computation of \GB{leading $0$'s} requires \GB{\texttt{clz}} instruction/algorithm 
-            \item Roughly as expensive as a \GB{\texttt{div}} instruction 
-        \end{itemize}
-    \end{block}
-\end{frame}

### 2024-02-27 Presentation/unbiased.tex
no black 0 in Java

@@
         \end{align*}} \only<4>{\begin{align*}\everymath{} % Allows specific coloring
-            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
+            \overbrace{\underbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
         \end{align*}} \only<5>{\begin{align*}\everymath{} % Allows specific coloring
-            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
+            \overbrace{\underbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
         \end{align*}} \only<6>{\begin{align*}\everymath{} % Allows specific coloring
-            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\substack{\text{mapped to} \\ 2^W - n - \RNW}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
+            \overbrace{\underbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\underbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}_{\substack{\text{mapped to} \\ 2^W - n - \RNW}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
         \end{align*}} \only<7->{\begin{align*}\everymath{} % Allows specific coloring
-            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\substack{\text{mapped to} \\ 2^W - n - \RNW}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\underbrace{\overbrace{\BB{0},1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}_{\substack{\text{mapped to} \\ 2^W - \RNW}}}
+            \overbrace{\underbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\underbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}_{\substack{\text{mapped to} \\ 2^W - n - \RNW}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\underbrace{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}_{\substack{\text{mapped to} \\ 2^W - \RNW}}}
         \end{align*}}
